---
title: "Liver Radiodensity & Chemotherapy Toxicity — Analysis Notebook"
author: "Noah Bos & Abdullah Alam"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    number_sections: true
    df_print: paged
    code_folding: show
---

This is the working analysis notebook. The canonical models are defined once in `R/main_analysis.R` and sourced both here and in the paper / slides. Use this notebook for:

- Inspecting raw model output
- Sensitivity analyses and alternative specifications
- Responding to feedback (with dated entries)
- Documenting decisions that don't make it into the paper

Anything experimental belongs here, **not** in `R/main_analysis.R`. If a sensitivity analysis becomes part of the canonical story, promote it.

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width  = 8,
  fig.height = 4.5
)

library(survival)
library(survminer)
library(dplyr)
library(tibble)
library(broom)
library(ggplot2)
library(forestmodel)
library(knitr)
library(kableExtra)
```

```{r load-analysis}
source("R/main_analysis.R")
```

# Data

## Cohort sizes

```{r}
tibble(
  Endpoint = c("CAP", "OXA", "CAPOX"),
  N        = c(nrow(cap), nrow(oxa), nrow(capox)),
  Events   = c(sum(cap$event), sum(oxa$event), sum(capox$event)),
  `Event rate (%)` = round(100 * c(mean(cap$event), mean(oxa$event), mean(capox$event)), 1)
) %>% kable()
```

## Hospital coverage

```{r}
df %>%
  group_by(hospital) %>%
  summarise(
    n_patients = n(),
    n_cap      = sum(!is.na(TIMT_CAP2)),
    n_oxa      = sum(!is.na(TIMT_OXA2)),
    n_capox    = sum(!is.na(TIMT_CAPOX2)),
    median_HU  = round(median(liverHU_manual, na.rm = TRUE), 1),
    .groups    = "drop"
  ) %>% arrange(desc(n_patients)) %>% kable()
```

# Models — raw output

Unformatted `summary()` for each Cox fit. Use this to spot anything unusual in standard errors, coefficients, or convergence.

## CAP

```{r}
summary(cox_cap)
```

## OXA

```{r}
summary(cox_oxa)
```

## CAPOX

```{r}
summary(cox_capox)
```

# Diagnostics

## Proportional hazards (Schoenfeld)

```{r}
ph_cap
ph_oxa
ph_capox
```

## Schoenfeld plots — CAPOX

```{r fig.height=8}
plot(ph_capox)
```

# Sensitivity Analyses

Add new sensitivity / alternative-specification chunks below. The canonical objects (`cap`, `oxa`, `capox`, `cox_*`, `df`) are already in scope from `R/main_analysis.R`. Promote a finding into the paper only after deciding it changes (or robustly confirms) the canonical conclusion.

<!--
## Example: tertiles of liver HU

```{r eval=FALSE}
capox$hu_tertile <- cut(capox$liverHU_manual,
                        breaks = quantile(capox$liverHU_manual, c(0, 1/3, 2/3, 1), na.rm = TRUE),
                        include.lowest = TRUE, labels = c("low", "mid", "high"))
cox_tertile <- coxph(Surv(time, event) ~ hu_tertile + age + sex + BBMI +
                       BSMOKER + doseCAP + doseOXA + strata(hospital), data = capox)
summary(cox_tertile)
```
-->

# Feedback Responses

Track feedback rounds here. Each entry: who, when, the comment, what we did.

<!--
## YYYY-MM-DD — reviewer / supervisor name

**Comment:** ...

**Response:** ...

**Changes:** files touched, decision rationale.
-->

# Decision Log

Non-obvious analytic decisions worth recording. Format: dated bullet, one or two sentences each.

<!--
- 2026-MM-DD — chose stratification over frailty model because the question is the within-hospital HU effect, not between-hospital variance.
-->
