---
title: "Liver Radiodensity & Chemotherapy Toxicity — Analysis Notebook"
author: "Noah Bos & Abdullah Alam"
date: "`r Sys.Date()`"
knit: (function(input, ...) rmarkdown::render(input, output_dir = "reports", output_file = "analysis.html"))
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    number_sections: true
    df_print: paged
    code_folding: show
---

This is the working analysis notebook. The canonical models are defined once in `R/main_analysis.R` and sourced both here and in the paper / slides. Use this notebook for:

- Inspecting raw model output
- Sensitivity analyses and alternative specifications
- Responding to feedback (with dated entries)
- Documenting decisions that don't make it into the paper

Anything experimental belongs here, **not** in `R/main_analysis.R`. If a sensitivity analysis becomes part of the canonical story, promote it.

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width  = 8,
  fig.height = 4.5
)

library(survival)
library(survminer)
library(dplyr)
library(tibble)
library(broom)
library(ggplot2)
library(forestmodel)
library(knitr)
library(kableExtra)
```

```{r load-analysis}
source("R/main_analysis.R")
```

# Data

## Cohort sizes

```{r}
tibble(
  Endpoint = c("CAP", "OXA", "CAPOX"),
  N        = c(nrow(cap), nrow(oxa), nrow(capox)),
  Events   = c(sum(cap$event), sum(oxa$event), sum(capox$event)),
  `Event rate (%)` = round(100 * c(mean(cap$event), mean(oxa$event), mean(capox$event)), 1)
) %>% kable()
```

## Hospital coverage

```{r}
df %>%
  group_by(hospital) %>%
  summarise(
    n_patients = n(),
    n_cap      = sum(!is.na(TIMT_CAP2)),
    n_oxa      = sum(!is.na(TIMT_OXA2)),
    n_capox    = sum(!is.na(TIMT_CAPOX2)),
    median_HU  = round(median(liverHU_manual, na.rm = TRUE), 1),
    .groups    = "drop"
  ) %>% arrange(desc(n_patients)) %>% kable()
```

# Models — raw output

Unformatted `summary()` for each Cox fit. Use this to spot anything unusual in standard errors, coefficients, or convergence.

## CAP

```{r}
summary(cox_cap)
```

## OXA

```{r}
summary(cox_oxa)
```

## CAPOX

```{r}
summary(cox_capox)
```

# Diagnostics

## Proportional hazards (Schoenfeld)

```{r}
ph_cap
ph_oxa
ph_capox
```

## Schoenfeld plots — CAPOX

```{r fig.height=8}
plot(ph_capox)
```

# Sensitivity Analyses

The canonical objects (`cap`, `oxa`, `capox`, `cox_*`, `df`) are in scope from `R/main_analysis.R`. Promote a finding into the paper only after deciding it changes (or robustly confirms) the canonical conclusion.

## Are hospital-level event rates more variable than chance? (Ophoff §3.2 query)

**The question.** Paper §3.2 claims toxicity rates differ across hospitals based on the observed range (~33% to ~80%). The reviewer asks whether this range could simply be **binomial sampling variation around a common rate**: even if every hospital had identical true toxicity probability, finite-sample noise alone would generate different observed proportions, with smaller hospitals fluctuating more. So we need to know whether what we see exceeds what sampling alone would produce.

**The test.** Build a hospital × event contingency table per endpoint and run **Pearson's chi-square test of independence**. The test statistic measures squared deviation between observed and expected counts under H₀: "all hospitals share the same toxicity rate." A large statistic (small p) means the observed table is unlikely under that null.

**Validity.** The chi-square asymptotic approximation needs expected cell counts ≥ 5. The smallest observed cell count across all hospital × event combinations in our data is 7, so the approximation is reliable here — Monte Carlo simulation isn't needed.

```{r hospital-rate-test}
hospital_rate_test <- function(data, name) {
  tab <- table(hospital = data$hospital, event = data$event)
  cs  <- chisq.test(tab)
  data.frame(
    Endpoint  = name,
    N         = sum(tab),
    Hospitals = nrow(tab),
    `Chi-sq`  = round(unname(cs$statistic), 2),
    df        = unname(cs$parameter),
    `p-value` = signif(cs$p.value, 3),
    check.names = FALSE
  )
}

bind_rows(
  hospital_rate_test(cap,   "CAP"),
  hospital_rate_test(oxa,   "OXA"),
  hospital_rate_test(capox, "CAPOX")
) %>% kable(caption = "Test of equal toxicity rates across hospitals (Pearson's chi-square)")
```

**How to read the p-value.**

- **p > 0.05** — observed range of hospital toxicity rates is *consistent with* binomial sampling noise around a common rate. We cannot reject H₀. The §3.2 wording must be softened: *"the observed range is consistent with sampling variation around a common rate"* rather than *"baseline risk differs between hospitals."*
- **0.01 < p < 0.05** — borderline evidence; mention the test result and let the reader judge.
- **p < 0.01** — strong evidence that hospitals differ in toxicity rates beyond sampling noise. The §3.2 claim stands as written.

**Either way, stratification is still justified.** Liver HU itself differs significantly across hospitals (ANOVA, §3.3), so the methodological motivation for stratifying — controlling for systematic measurement and case-mix differences — does not depend on this test.

```{r hospital-rate-table}
event_long <- bind_rows(
  transmute(df, hospital, endpoint = "CAP",   event = TIMT_CAP2),
  transmute(df, hospital, endpoint = "OXA",   event = TIMT_OXA2),
  transmute(df, hospital, endpoint = "CAPOX", event = TIMT_CAPOX2)
) %>% filter(!is.na(event))

event_long %>%
  group_by(endpoint, hospital) %>%
  summarise(N = n(), Events = sum(event == 1),
            `Rate (%)` = round(100 * mean(event == 1), 1), .groups = "drop") %>%
  arrange(endpoint, desc(N)) %>%
  kable(caption = "Hospital × endpoint event counts (raw, for inspection)")
```

**Interpretation:** if `p > 0.05` the observed range is consistent with sampling noise around a common rate; the descriptive claim in §3.2 needs softening to *"observed rates are consistent with sampling variation"* rather than *"baseline risk differs between hospitals"*. The methodological case for stratification still holds (HU **does** differ across hospitals — see ANOVA in §3.3).

## Stratified vs. unstratified Cox (Ophoff sub-question 2)

To answer sub-question 2 ("how does accounting for hospital affect the estimated HU association"), refit each Cox model **without** `strata(hospital)` and place the HU hazard ratio side by side with the stratified version.

```{r unstratified-fits}
cox_cap_unstrat <- coxph(
  Surv(time, event) ~ liverHU_manual + age + sex + BBMI +
    BSMOKER + starting_dose, data = cap)

cox_oxa_unstrat <- coxph(
  Surv(time, event) ~ liverHU_manual + age + sex + BBMI +
    BSMOKER + starting_dose, data = oxa)

cox_capox_unstrat <- coxph(
  Surv(time, event) ~ liverHU_manual + age + sex + BBMI +
    BSMOKER + doseCAP + doseOXA, data = capox)
```

```{r strat-vs-unstrat}
hu_row <- function(model, label, model_type) {
  t <- tidy(model, exponentiate = TRUE, conf.int = TRUE) %>%
    filter(term == "liverHU_manual")
  data.frame(
    Endpoint = label, Model = model_type,
    HR = round(t$estimate, 3),
    `95% CI` = sprintf("%.3f–%.3f", t$conf.low, t$conf.high),
    `p-value` = signif(t$p.value, 3),
    check.names = FALSE
  )
}

bind_rows(
  hu_row(cox_cap,           "CAP",   "Stratified"),
  hu_row(cox_cap_unstrat,   "CAP",   "Unstratified"),
  hu_row(cox_oxa,           "OXA",   "Stratified"),
  hu_row(cox_oxa_unstrat,   "OXA",   "Unstratified"),
  hu_row(cox_capox,         "CAPOX", "Stratified"),
  hu_row(cox_capox_unstrat, "CAPOX", "Unstratified")
) %>% kable(caption = "Liver HU hazard ratio (per 1 HU): stratified vs. unstratified")
```

**Interpretation:** any meaningful shift between the two columns is the contribution of hospital-level confounding. If the estimates are similar, hospital adjustment isn't doing much; if they differ, the unstratified estimate is biased by between-hospital differences in HU and toxicity rates.

## Rescaled liver HU effect (Ophoff: per-1-HU is tiny)

The reviewer notes that HR per 1 HU looks trivially close to 1, and suggests presenting per 10 HU and/or per SD. We compute both for the stratified models.

```{r rescaled-hu}
rescale_hu <- function(model, label, data, scales = c("per 1 HU", "per 10 HU", "per SD")) {
  cf <- coef(model)["liverHU_manual"]
  se <- sqrt(vcov(model)["liverHU_manual", "liverHU_manual"])
  sd_hu <- sd(data$liverHU_manual, na.rm = TRUE)

  k <- c(`per 1 HU` = 1, `per 10 HU` = 10, `per SD` = sd_hu)[scales]

  data.frame(
    Endpoint = label,
    Scale    = scales,
    `HR`     = round(exp(cf * k), 3),
    `95% CI` = sprintf("%.3f–%.3f", exp((cf - 1.96 * se) * k), exp((cf + 1.96 * se) * k)),
    check.names = FALSE
  )
}

bind_rows(
  rescale_hu(cox_cap,   "CAP",   cap),
  rescale_hu(cox_oxa,   "OXA",   oxa),
  rescale_hu(cox_capox, "CAPOX", capox)
) %>% kable(caption = "Liver HU hazard ratio at three scalings (stratified models)")
```

**Interpretation:** the per-10-HU and per-SD columns expose whether the effect is clinically meaningful at realistic exposure differences. SD of liver HU in this cohort is roughly 12–14 — a per-SD HR is the most directly interpretable.

# Feedback Responses

## 2026-05-09 — Ophoff, Written Report v3

**Comments and responses:**

| # | Comment | Response | Where |
|---|---|---|---|
| 1 | Sub-question 3 is vague | Rewrite | `paper.Rmd` Introduction |
| 2 | "Frailty model" jargon | Drop or define | `paper.Rmd` Methods |
| 3 | HR per 1 HU looks tiny — rescale | Present per 10 HU and per SD | This notebook §"Rescaled liver HU effect"; promote chosen scaling to `paper.Rmd` Tables 4 & 5 |
| 4 | §3.2 hospital differences could be sampling noise | Chi-square test added | This notebook §"Are hospital-level event rates more variable than chance?"; rephrase paper §3.2 based on result |
| 5 | Sub-question 2 not answered — compare stratified vs. unstratified | Side-by-side comparison added | This notebook §"Stratified vs. unstratified Cox"; promote table to `paper.Rmd` |
| 6 | Sub-question 3 not answered | Address in Discussion | `paper.Rmd` Discussion |
| 7 | Cite R version + packages | Add citation | `paper.Rmd` Methods (Software) |
| 8 | Reconsider clinical-relevance claim | Update after rescaling | `paper.Rmd` end of Results 3.4 |
| 9 | Sex "protective" wording awkward | Rephrase to name protected group | `paper.Rmd` Results §3.4 commentary |
| 10 | p.3: "that they correctly handle censoring" → "the correct handling of censoring" | Fix | `paper.Rmd` Methods |
| 11 | Tables 3 / 4 / 5: digits, NAs, format `1` → `1.00` | Fix table formatting | `paper.Rmd` Results |
| 12 | Tables/figures should be more standalone | Improve captions | `paper.Rmd` throughout |

# Decision Log

<!--
- 2026-MM-DD — chose stratification over frailty model because the question is the within-hospital HU effect, not between-hospital variance.
-->
