---
title: "Liver Radiodensity and Chemotherapy Toxicity"
subtitle: "Survival Analysis in a Dutch Multi-centre Cohort"
author: "Noah Bos & Abdullah Alam"
institute: "COLON Study · Wageningen University Research"
date: "`r Sys.Date()`"
knit: (function(input, ...) rmarkdown::render(input, output_dir = "reports", output_file = "liverfat_survival_slides_v2.pdf"))
classoption: "aspectratio=169"
output:
  beamer_presentation:
    theme: "metropolis"
    keep_tex: false
    slide_level: 2
header-includes:
  - \usepackage{booktabs}
  - \usepackage{amsmath}
  - \metroset{block=fill}
---

```{r install-packages, include=FALSE}
required_packages <- c(
  "survival", "survminer", "dplyr", "knitr",
  "broom", "ggplot2", "forestmodel", "kableExtra", "tibble"
)
missing <- required_packages[!sapply(required_packages, requireNamespace, quietly = TRUE)]
if (length(missing) > 0) install.packages(missing)
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.width  = 11,
  fig.height = 3.8
)

library(survival)
library(survminer)
library(dplyr)
library(tibble)
library(knitr)
library(kableExtra)
library(broom)
library(ggplot2)
library(forestmodel)
```

```{r data-prep, include=FALSE}
load("data/Dataset_liverfat.RData")
df <- dataset_leiden

make_endpoint <- function(data, endpoint) {
  if (endpoint == "CAP") {
    event <- data$TIMT_CAP2
    time  <- data$cycle_till_timt
    dose  <- data$starting_doseCAP
  }
  if (endpoint == "OXA") {
    event <- data$TIMT_OXA2
    time  <- data$cycle_till_timtOXA
    dose  <- data$starting_doseOXA
  }
  if (endpoint == "CAPOX") {
    event   <- data$TIMT_CAPOX2
    time    <- data$cycle_till_timtCAPOX
    doseCAP <- data$starting_doseCAP
    doseOXA <- data$starting_doseOXA
  }

  out <- data.frame(
    time, event,
    liverHU_manual = data$liverHU_manual,
    age      = data$age,
    sex      = data$sex,
    BBMI     = data$BBMI,
    BSMOKER  = data$BSMOKER,
    hospital = data$hospital
  )

  if (endpoint == "CAP" || endpoint == "OXA") out$starting_dose <- dose
  if (endpoint == "CAPOX") { out$doseCAP <- doseCAP; out$doseOXA <- doseOXA }

  out[complete.cases(out), ]
}

cap   <- make_endpoint(df, "CAP")
oxa   <- make_endpoint(df, "OXA")
capox <- make_endpoint(df, "CAPOX")

cox_cap <- coxph(
  Surv(time, event) ~ liverHU_manual + age + sex + BBMI +
    BSMOKER + starting_dose + strata(hospital), data = cap
)
cox_oxa <- coxph(
  Surv(time, event) ~ liverHU_manual + age + sex + BBMI +
    BSMOKER + starting_dose + strata(hospital), data = oxa
)
cox_capox <- coxph(
  Surv(time, event) ~ liverHU_manual + age + sex + BBMI +
    BSMOKER + doseCAP + doseOXA + strata(hospital), data = capox
)

hu_cap   <- tidy(cox_cap,   exponentiate = TRUE, conf.int = TRUE) %>% filter(term == "liverHU_manual")
hu_oxa   <- tidy(cox_oxa,   exponentiate = TRUE, conf.int = TRUE) %>% filter(term == "liverHU_manual")
hu_capox <- tidy(cox_capox, exponentiate = TRUE, conf.int = TRUE) %>% filter(term == "liverHU_manual")

hu_range    <- range(capox$liverHU_manual, na.rm = TRUE)
hu_iqr      <- IQR(capox$liverHU_manual, na.rm = TRUE)
hu_q25      <- quantile(capox$liverHU_manual, 0.25, na.rm = TRUE)
hu_q75      <- quantile(capox$liverHU_manual, 0.75, na.rm = TRUE)

hr_iqr_capox   <- round(hu_capox$estimate ^ hu_iqr,   2)
hr_range_capox <- round(hu_capox$estimate ^ diff(hu_range), 2)

cap_hosp_rates <- df %>%
  filter(!is.na(TIMT_CAP2)) %>%
  group_by(hospital) %>%
  summarise(rate = mean(TIMT_CAP2 == 1, na.rm = TRUE), .groups = "drop")
```

## Background \& Research Question

**Colorectal cancer (stage II/III)** is treated with capecitabine (CAP), oxaliplatin (OXA), or both (CAPOX). These cause toxicities — hand-foot syndrome, neuropathy — that force dose reductions or early discontinuation.

**Candidate predictor:** Liver fat measured as radiodensity in **Hounsfield Units (HU)** on CT scans at diagnosis. Lower HU = more fat = potentially impaired drug metabolism.

\vspace{4pt}

\begin{block}{Research question}
Does liver radiodensity predict \textit{when} a patient experiences dose-limiting toxicity — after accounting for differences across the 11 hospitals in the COLON cohort (Wageningen University Research)?
\end{block}

## Study Design

```{r summary-table}
endpoint_summary <- tibble(
  Endpoint         = c("CAP", "OXA", "CAPOX"),
  N                = c(nrow(cap), nrow(oxa), nrow(capox)),
  Events           = c(sum(cap$event), sum(oxa$event), sum(capox$event)),
  Censored         = c(sum(!cap$event), sum(!oxa$event), sum(!capox$event)),
  `Event rate (%)` = round(c(mean(cap$event), mean(oxa$event), mean(capox$event)) * 100, 1)
)
kable(endpoint_summary, format = "latex", booktabs = TRUE, linesep = "") %>%
  kable_styling(font_size = 11, latex_options = "hold_position")
```

\vspace{4pt}

**Outcome:** Chemotherapy cycles until dose-limiting toxicity (or censored if none) \
**Exposure:** Liver HU — CT scan at diagnosis \
**Adjusted for:** Age, sex, BMI, smoking, starting dose \
**Three separate models** — one per treatment

## Survival Analysis: What Is It Used For?

Survival analysis models **time until an event** — whenever you care about *when*, not just *if*.

\vspace{4pt}

:::::: {.columns}

::: {.column width="48%"}

**Healthcare & life sciences**

- Time to disease recurrence after surgery
- Time to death or hospital readmission
- Time to side-effects during treatment

\vspace{6pt}

**Business & consulting**

- Customer churn: when does a subscriber cancel?
- Time to first purchase after sign-up
- Employee attrition: when do staff leave?

:::

::: {.column width="48%"}

**Finance & risk**

- Time to loan default
- Time to insurance claim

\vspace{6pt}

**Engineering & operations**

- Machine failure / predictive maintenance
- Time to system outage

\vspace{6pt}

**Digital / product**

- Time to user conversion or drop-off
- A/B test: which variant converts faster?

:::

::::::

## Why Survival Analysis? Censoring \& Core Concepts

:::::: {.columns}

::: {.column width="48%"}

**The censoring problem:** Not every patient experiences toxicity. A patient finishing 8 cycles event-free is not a failure — they tell us *"at least 8 cycles without toxicity."* Logistic regression discards this information entirely.

\vspace{6pt}

Any model that ignores censored observations will **overestimate** risk by treating non-events as missing data.

:::

::: {.column width="48%"}

**Survival function** — probability of remaining event-free beyond time $t$:
$$S(t) = P(T > t)$$

**Hazard function** — instantaneous event rate at time $t$, given event-free so far:
$$h(t) = \lim_{\Delta t \to 0} \frac{P(t \leq T < t {+} \Delta t \mid T \geq t)}{\Delta t}$$

They carry the same information:
$$S(t) = \exp\!\left(-\int_0^t h(u)\, du\right)$$

:::

::::::

## The Cox Model

:::::: {.columns}

::: {.column width="52%"}

```{r cox-curve, fig.height=3.4, fig.width=6.2}
set.seed(42)
n <- 40
dummy <- data.frame(
  time  = c(rexp(n, rate = 0.30), rexp(n, rate = 0.12)),
  event = rbinom(2 * n, 1, prob = 0.85),
  group = rep(c("Group A (higher risk)", "Group B (lower risk)"), each = n)
)
dummy$time <- pmax(ceiling(dummy$time), 1)

fit_dummy <- survfit(Surv(time, event) ~ group, data = dummy)
ggsurvplot(
  fit_dummy,
  palette      = c("#D55E00", "#0072B2"),
  conf.int     = FALSE,
  legend.title = "",
  legend       = "bottom",
  xlab         = "Time",
  ylab         = "S(t)  —  Prob. event-free",
  title        = "Kaplan-Meier curves (illustrative)",
  ggtheme      = theme_minimal(base_size = 11)
)
```

:::

::: {.column width="44%"}

**Cox model** — regression on the hazard:

$$h(t \mid \mathbf{x}) = h_0(t) \cdot e^{\,\beta_1 x_1 + \cdots + \beta_p x_p}$$

\vspace{4pt}

$h_0(t)$ = baseline hazard, left unspecified

\vspace{8pt}

**Hazard ratio:** $\text{HR}_j = e^{\hat\beta_j}$

\vspace{4pt}

HR $= 1.20 \Rightarrow$ 20\% higher risk at every cycle per one-unit increase in $x_j$

\vspace{4pt}

HR $< 1 \Rightarrow$ protective; HR $> 1 \Rightarrow$ harmful

:::

::::::

## Hospital Heterogeneity

```{r hosp-plot, fig.height=3.5}
hosp_anova <- anova(lm(liverHU_manual ~ hospital, data = df))
f_val <- round(hosp_anova$`F value`[1], 1)
p_val <- format.pval(hosp_anova$`Pr(>F)`[1], digits = 2, eps = 0.001)

ggplot(df, aes(x = reorder(hospital, liverHU_manual, FUN = median, na.rm = TRUE),
               y = liverHU_manual)) +
  geom_boxplot(fill = "#56B4E9", alpha = 0.7, outlier.size = 0.8) +
  labs(
    x        = "Hospital (ordered by median HU)",
    y        = "Liver radiodensity (HU)",
    title    = "Liver HU differs significantly across hospitals",
    subtitle = paste0("One-way ANOVA: F = ", f_val, ", p ", p_val,
                      "  —  CT scanner differences and patient mix")
  ) +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```

CAP toxicity rates range from `r round(min(cap_hosp_rates$rate)*100)`\% to `r round(max(cap_hosp_rates$rate)*100)`\% across hospitals. Both exposure and outcome differ by hospital — a textbook confounding situation.

## Stratified Cox Model

Instead of adding hospital as a covariate (which forces one shared hazard curve), we **stratify**:

$$h_k(t \mid \mathbf{x}) = \underbrace{h_{0k}(t)}_{\text{hospital-specific baseline}} \cdot \; e^{\,\boldsymbol{\beta}^\top \mathbf{x}}$$

Hospital $k$ gets its **own** $h_{0k}(t)$, absorbing all between-hospital differences. The liver HU effect $\boldsymbol{\beta}$ is estimated from **within-hospital comparisons only**.

\vspace{6pt}

\begin{block}{Why this matters}
Without stratification, the liver HU estimate is contaminated by between-hospital differences in both the HU measurement and toxicity rates. Stratification removes this confounding entirely.
\end{block}

## Results

```{r hu-hr-table}
hr_hu <- bind_rows(
  tidy(cox_cap,   exponentiate = TRUE, conf.int = TRUE) %>%
    filter(term == "liverHU_manual") %>% mutate(Endpoint = "CAP"),
  tidy(cox_oxa,   exponentiate = TRUE, conf.int = TRUE) %>%
    filter(term == "liverHU_manual") %>% mutate(Endpoint = "OXA"),
  tidy(cox_capox, exponentiate = TRUE, conf.int = TRUE) %>%
    filter(term == "liverHU_manual") %>% mutate(Endpoint = "CAPOX")
) %>%
  mutate(
    `HR (95% CI)` = paste0(round(estimate, 2), " (",
                            round(conf.low, 2), "–", round(conf.high, 2), ")"),
    `p-value`     = round(p.value, 3),
    Interpretation = dplyr::case_when(
      p.value < 0.05 ~ "Significant",
      p.value < 0.10 ~ paste0("Borderline (p = ", round(p.value, 3), ")"),
      TRUE           ~ "Not significant"
    )
  ) %>%
  select(Endpoint, `HR (95% CI)`, `p-value`, Interpretation)

kable(hr_hu, format = "latex", booktabs = TRUE, linesep = "",
      caption = "Liver radiodensity effect (per 1 HU) on time to toxicity — stratified Cox models") %>%
  kable_styling(font_size = 11, latex_options = "hold_position")
```

\vspace{4pt}

HR = 1.01 per HU looks small, but liver HU ranges from `r round(hu_range[1])` to `r round(hu_range[2])` in this cohort. A patient with high liver fat vs. one with little fat can differ by `r round(diff(hu_range))` HU — where the hazard compounds to HR $\approx$ **`r hr_range_capox`**, a `r round((hr_range_capox-1)*100)`\% difference in toxicity risk.

## CAPOX: Full Model Results

```{r forest-capox, fig.height=4.0}
forest_model(cox_capox)
```

## Conclusion

\begin{block}{Answer}
After stratifying by hospital, liver radiodensity shows a statistically significant association with CAPOX toxicity timing. Per HU the effect is small, but across the observed HU range the cumulative hazard difference is substantial.
\end{block}

\vspace{6pt}

**Three take-aways:**

1. **Hospital heterogeneity is real** — liver HU and toxicity rates differ across all 11 hospitals. Ignoring it biases the results.

2. **Stratification beats covariate adjustment** when hospitals differ in their entire baseline hazard curve, not just in a single HR.

3. **Scale matters for interpretation** — HR = `r round(hu_capox$estimate, 2)` per HU sounds trivial, but the full observed HU range spans `r round(diff(hu_range))` units, giving a compounded HR of `r hr_range_capox`. Always put effect sizes in the context of the variable's real-world range.
