---
title: "Hospital-Level Variation in Liver Radiodensity and Chemotherapy-Induced Toxicities in Colon Cancer Patients"
author: "Abdullah Abdullah & Noah Bos"
date: "`r Sys.Date()`"
output:
  pdf_document:
    latex_engine: xelatex
    number_sections: true
    keep_tex: false
    fig_caption: true
header-includes:
  - \usepackage{booktabs}
  - \usepackage{longtable}
  - \usepackage{float}
  - \usepackage{setspace}
  - \usepackage{parskip}
  - \usepackage{caption}
  - \usepackage{titling}
  - \onehalfspacing
  - \floatplacement{figure}{H}
  - \floatplacement{table}{H}
  - \renewcommand{\maketitle}{}
fontsize: 11pt
geometry: margin=2.5cm
---

```{r install-packages, include=FALSE}
required_packages <- c(
  "survival", "survminer", "dplyr", "knitr",
  "broom", "ggplot2", "forestmodel", "kableExtra", "tibble"
)
missing <- required_packages[!sapply(required_packages, requireNamespace, quietly = TRUE)]
if (length(missing) > 0) install.packages(missing)
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.pos = "H",
  fig.width  = 6,
  fig.height = 4
)

library(survival)
library(survminer)
library(dplyr)
library(tibble)
library(knitr)
library(kableExtra)
library(broom)
library(ggplot2)
library(forestmodel)
```

```{r data-prep, include=FALSE}
load("data/Dataset_liverfat.RData")
df <- dataset_leiden

make_endpoint <- function(data, endpoint) {
  if (endpoint == "CAP") {
    event <- data$TIMT_CAP2
    time  <- data$cycle_till_timt
    dose  <- data$starting_doseCAP
  }
  if (endpoint == "OXA") {
    event <- data$TIMT_OXA2
    time  <- data$cycle_till_timtOXA
    dose  <- data$starting_doseOXA
  }
  if (endpoint == "CAPOX") {
    event   <- data$TIMT_CAPOX2
    time    <- data$cycle_till_timtCAPOX
    doseCAP <- data$starting_doseCAP
    doseOXA <- data$starting_doseOXA
  }

  out <- data.frame(
    time, event,
    liverHU_manual = data$liverHU_manual,
    age      = data$age,
    sex      = data$sex,
    BBMI     = data$BBMI,
    BSMOKER  = data$BSMOKER,
    hospital = data$hospital
  )

  if (endpoint == "CAP" || endpoint == "OXA") out$starting_dose <- dose
  if (endpoint == "CAPOX") { out$doseCAP <- doseCAP; out$doseOXA <- doseOXA }

  out[complete.cases(out), ]
}

cap   <- make_endpoint(df, "CAP")
oxa   <- make_endpoint(df, "OXA")
capox <- make_endpoint(df, "CAPOX")
```

```{r fit-models, include=FALSE}
cox_cap <- coxph(
  Surv(time, event) ~ liverHU_manual + age + sex + BBMI +
    BSMOKER + starting_dose + strata(hospital),
  data = cap
)

cox_oxa <- coxph(
  Surv(time, event) ~ liverHU_manual + age + sex + BBMI +
    BSMOKER + starting_dose + strata(hospital),
  data = oxa
)

cox_capox <- coxph(
  Surv(time, event) ~ liverHU_manual + age + sex + BBMI +
    BSMOKER + doseCAP + doseOXA + strata(hospital),
  data = capox
)

ph_cap   <- cox.zph(cox_cap)
ph_oxa   <- cox.zph(cox_oxa)
ph_capox <- cox.zph(cox_capox)
```

```{=latex}
\begin{titlepage}
\centering
\vspace*{3cm}

{\Large \textbf{Hospital-Level Variation in Liver Radiodensity and\\[0.6em]
Chemotherapy-Induced Toxicities in\\[0.6em]
Colon Cancer Patients}}

\vspace{2cm}

{\large Abdullah Abdullah \& Noah Bos}

\vspace{1cm}

{\normalsize \today}

\vspace{2cm}

{\normalsize Statistical Consulting Project\\[0.3em]
COLON Cohort Study\\[0.3em]
Leiden University \& Wageningen University \& Research}

\vfill
\end{titlepage}
\newpage
\setcounter{page}{1}
```

# Introduction

Capecitabine and oxaliplatin are commonly used chemotherapeutic agents in the treatment of stage II and stage III colorectal cancer, either as monotherapy or in combination (CAPOX regimen). Although effective, both agents are associated with clinically relevant toxicities that may lead to dose reductions, treatment delays, or early discontinuation. Capecitabine-related toxicities frequently include hand-foot syndrome and gastrointestinal side-effects (Scheithauer et al., 2003; Iacovelli et al., 2014). In contrast, oxaliplatin is particularly associated with peripheral neuropathy, a dose-limiting toxicity that develops in a large proportion of treated patients and may persist over time (Renting et al., 2024). It is therefore critical to identify factors that influence treatment-related toxicities, and this remains an important clinical and research priority.

Recent analyses within the COLON cohort, a prospective multicentre cohort study conducted in the Netherlands, showed that circulating folic acid and folic acid supplement use were associated with an increased risk of capecitabine-induced toxicities among patients with colon or rectal cancer (Kok et al., 2024). Additional work within this cohort has examined nutritional determinants of chronic oxaliplatin-induced toxicities. These findings suggest that metabolic and nutritional factors may predispose patients to chemotherapy-induced toxicities, although causal relations cannot be established within observational designs.

In the current project, we focus on liver fat accumulation, measured as liver radiodensity on computed tomography (CT) scans taken at the time of diagnosis, as a potential determinant of chemotherapy-induced side-effects. The data originate from the prospective COLON study, in which patients newly diagnosed with colorectal cancer were recruited from 11 hospitals in the Netherlands (Winkels et al., 2014). While inclusion of patients from different hospitals improves generalisability, it also introduces potential heterogeneity. Differences in CT scanner characteristics, imaging protocols, and calibration procedures across hospitals may lead to systematic variation in measured liver radiodensity. In addition, differences in patient characteristics across hospitals (such as age, tumour stage, or general health status) or differences in clinical practices across hospitals may influence the occurrence of treatment-related toxicities.

If hospital-level differences are present, they may affect the estimated association between liver radiodensity and risk of chemotherapy-induced toxicities. Ignoring such variation could lead to biased estimates or misleadingly precise confidence intervals. Therefore, it is important to assess whether hospital-level effects exist and to determine how they influence statistical inference. The primary aim of this paper is to evaluate how hospital-level variation influences the estimated risk and timing of chemotherapy-induced toxicities in relation to liver radiodensity at diagnosis within a multicentre observational cohort. Specifically, we address the following main research question:

\bigskip
\noindent\textbf{RQ:} \textit{How does hospital-level variation influence the estimated association between liver radiodensity and time to chemotherapy-induced toxicity?}
\bigskip

The study specifically answers the following sub-questions:

- Are there systematic differences between hospitals in liver radiodensity measurements and/or toxicity outcomes?
- How does accounting for hospital-level variation affect the estimated association between liver radiodensity and toxicity?
- How should hospital-level heterogeneity be incorporated when interpreting the findings?

# Methods

This section describes the study design, variables, and statistical approach used to investigate the association between liver radiodensity and time to chemotherapy-induced toxicity. Particular attention is given to the multicentre nature of the data and the potential role of hospital-level variation.

## Participants and Study Design

The data originate from a prospective multicentre cohort study conducted in the Netherlands, in which patients diagnosed with colorectal cancer were recruited from multiple hospitals and followed during chemotherapy treatment. The study has an observational design, and no intervention was assigned as part of this analysis. Patients were included if they received chemotherapy involving capecitabine, oxaliplatin, or a combination of both (CAPOX regimen), and if computed tomography (CT) scans at the time of diagnosis were available. Because not all patients received the same treatment, separate analytic samples were defined for each toxicity endpoint. Patients were excluded from a given analysis if they were not at risk for the corresponding toxicity or if information on time to toxicity or event occurrence was missing.

## Measurements

**Outcome: Time to Dose-Limiting Toxicity.** The primary outcome is time to first occurrence of dose-limiting toxicity during chemotherapy treatment. Time was operationalised as the number of chemotherapy cycles completed before the onset of toxicity. For each endpoint, the outcome consists of a time variable and an event indicator specifying whether toxicity occurred. Patients who did not experience toxicity during the observed treatment period were treated as censored at their last completed cycle. Three toxicity outcomes were considered: toxicity associated with capecitabine (CAP), toxicity associated with oxaliplatin (OXA), and toxicity associated with the combined CAPOX regimen. Dose-limiting toxicity was defined as any toxicity event leading to a dose reduction, treatment delay, or premature discontinuation of chemotherapy. Each outcome defines a separate risk set, as only patients receiving the corresponding treatment are at risk for that specific toxicity.

**Exposure: Liver Radiodensity.** The main exposure is liver radiodensity measured on CT scans at the time of diagnosis. Radiodensity is expressed in Hounsfield Units (HU), with lower values indicating higher liver fat content. Liver radiodensity was analysed as a continuous variable.

**Covariates.** The analyses adjust for a set of patient and treatment characteristics that may influence the risk of chemotherapy-induced toxicity, including age at diagnosis, sex, body mass index, smoking status, and starting dose of chemotherapy. Starting dose was defined in a treatment-specific manner depending on the toxicity endpoint considered: the capecitabine starting dose was used for the CAP endpoint, the oxaliplatin starting dose for the OXA endpoint, and both doses were included simultaneously for the CAPOX endpoint.

**Hospital.** Patients were recruited from multiple hospitals. Hospital was treated as a grouping variable to account for potential differences in imaging procedures, patient characteristics, and clinical practices across centres.

## Statistical Analysis

**Descriptive Analysis.** Descriptive statistics were used to summarise the study population and to explore the distribution of toxicity outcomes across endpoints and hospitals. Event rates were calculated for each toxicity endpoint, and variation in toxicity occurrence across hospitals was examined.

Differences in liver radiodensity across hospitals were assessed using one-way analysis of variance (ANOVA). This analysis evaluates whether mean radiodensity values differ between hospitals and provides insight into potential measurement heterogeneity across centres.

**Survival Analysis.** The association between liver radiodensity and time to dose-limiting toxicity was analysed using Cox proportional hazards regression models. Separate models were fitted for each toxicity endpoint (CAP, OXA, and CAPOX), using endpoint-specific datasets that include only patients at risk for the respective toxicity. The primary exposure in all models was liver radiodensity. Models were adjusted for age, sex, body mass index, smoking status, and treatment-specific starting dose. To account for hospital-level differences, stratified Cox models were used. Stratification was preferred over a frailty model because the primary aim was to estimate the within-hospital effect of liver radiodensity rather than to quantify between-hospital variability as a parameter of interest. Stratification makes no distributional assumptions about hospital-level effects and is therefore a more conservative and robust approach in this setting. This approach allows the baseline hazard to vary across hospitals while estimating a common effect of liver radiodensity across centres.

Cox proportional hazards models estimate the association between one or more variables and the time until an event occurs, in this case dose-limiting toxicity. The model is based on the concept of the hazard: the rate at which the event occurs at any given point in time, among patients who have not yet experienced it. The main output is the hazard ratio (HR). A value of 1.0 indicates no association; values above 1.0 indicate a higher rate of toxicity and values below 1.0 a lower rate. A key advantage of Cox models is that they correctly handle censoring. Patients who complete treatment without experiencing toxicity, or who leave follow-up early, are included in the analysis rather than excluded. Their observation time contributes to the model up to the point they are last observed.

Stratification by hospital means that each hospital is given its own baseline hazard function, estimated from the patients within that hospital only. The effect of liver radiodensity is then estimated as a single common effect across hospitals, based entirely on within-hospital comparisons. This differs from including hospital as a regular covariate, which assumes the baseline hazard has the same shape in every hospital. Stratification makes no such assumption and is therefore more flexible. The trade-off is that no single estimate of the hospital effect is produced, but since the aim of this analysis was to estimate the effect of liver radiodensity rather than to describe hospital-level differences, this is acceptable.

**Model Assumptions.** The proportional hazards assumption was assessed using Schoenfeld residual tests for all three models. This assumption implies that the effect of liver radiodensity on toxicity risk remains constant over time. Missingness in covariate data was assessed prior to analysis. No imputation was performed, as the proportion of missing values was low and unlikely to meaningfully affect the results.

**Software.** All analyses were conducted using R (version `r R.version$major`.`r R.version$minor`), with survival models estimated using the \texttt{survival} package.

# Results

## Sample Characteristics and Event Rates

```{r compute-n, include=FALSE}
n_cap_raw   <- sum(!is.na(df$TIMT_CAP2))
n_oxa_raw   <- sum(!is.na(df$TIMT_OXA2))
n_capox_raw <- sum(!is.na(df$TIMT_CAPOX2))
```

The analysis included patients receiving capecitabine (CAP), oxaliplatin (OXA), or a combination of both (CAPOX). Because not all patients received each treatment, separate datasets were constructed for each toxicity endpoint. The number of patients eligible for each endpoint (non-missing event indicator) was `r n_cap_raw`, `r n_oxa_raw`, and `r n_capox_raw` for CAP, OXA, and CAPOX respectively. After restricting to patients with complete covariate information required for the Cox models, the analytic samples comprised `r nrow(cap)`, `r nrow(oxa)`, and `r nrow(capox)` patients respectively. Table 1 presents the event counts for each analytic sample.

```{r table1}
tbl1 <- tibble(
  Endpoint  = c("CAP", "OXA", "CAPOX"),
  N         = c(nrow(cap),           nrow(oxa),           nrow(capox)),
  Events    = c(sum(cap$event),      sum(oxa$event),      sum(capox$event)),
  Censored  = c(sum(cap$event == 0), sum(oxa$event == 0), sum(capox$event == 0)),
  `Event rate (%)` = round(100 * c(
    sum(cap$event)   / nrow(cap),
    sum(oxa$event)   / nrow(oxa),
    sum(capox$event) / nrow(capox)), 1)
)

kable(tbl1, booktabs = TRUE,
      caption = "Observed toxicity rates by endpoint (complete-case analytic sample)") %>%
  kable_styling(latex_options = c("hold_position"), font_size = 10)
```

Across the three endpoints, dose-limiting toxicity was a common outcome. The proportion was highest for oxaliplatin-related toxicity. The number of observed events was sufficient to support time-to-event analyses for all endpoints.

## Variation in Toxicity Across Hospitals

Substantial variation in toxicity rates was observed across hospitals. The proportion of patients experiencing toxicity ranged from approximately one-third to as high as four-fifths across centres, indicating that baseline risk of toxicity differs between hospitals. Table 2 shows the distribution of CAP toxicity outcomes by hospital.

```{r table2}
tbl2 <- df %>%
  filter(!is.na(TIMT_CAP2)) %>%
  group_by(Hospital = hospital) %>%
  summarise(
    N          = n(),
    Events     = sum(TIMT_CAP2 == 1),
    `Event (%)` = round(100 * Events / n(), 1),
    Censored   = sum(TIMT_CAP2 == 0),
    .groups    = "drop"
  ) %>%
  arrange(desc(N))

kable(tbl2, booktabs = TRUE,
      caption = "Distribution of CAP toxicity outcomes by hospital") %>%
  kable_styling(latex_options = c("hold_position"), font_size = 10)
```

These differences may reflect variation in patient characteristics, treatment practices, or reporting procedures across hospitals. The presence of such variation highlights the importance of accounting for hospital-level differences in the statistical analysis.

## Variation in Liver Radiodensity Across Hospitals

Visual inspection of liver radiodensity distributions indicated differences across hospitals (Figure 1). This was confirmed by a one-way ANOVA, which showed that mean liver radiodensity differed significantly between hospitals. Table 3 presents the ANOVA results.

```{r figure1, fig.cap="Distribution of liver radiodensity (HU) by hospital, ordered by median HU.", fig.height=3.5}
ggplot(df, aes(x = reorder(hospital, liverHU_manual, median),
               y = liverHU_manual)) +
  geom_boxplot(fill = "#AEC6E8", color = "#2166AC") +
  labs(x = "Hospital (ordered by median HU)", y = "Liver radiodensity (HU)") +
  theme_minimal(base_size = 10)
```

```{r table3}
hospital_anova <- anova(lm(liverHU_manual ~ hospital, data = df))

kable(as.data.frame(hospital_anova), booktabs = TRUE, digits = 3,
      caption = "One-way ANOVA: liver radiodensity across hospitals") %>%
  kable_styling(latex_options = c("hold_position"), font_size = 10)
```

These findings indicate that liver radiodensity measurements are not directly comparable across hospitals and may be influenced by differences in imaging procedures or patient populations. This suggests that hospital may act as a confounding factor in the association between liver radiodensity and toxicity.

## Time to Toxicity

Kaplan-Meier curves were used to describe the unadjusted time to dose-limiting toxicity across chemotherapy cycles (Figure 2). Across all endpoints, the probability of remaining free from toxicity declined rapidly during the early treatment cycles, indicating that most toxicities occurred early in treatment. In later cycles, estimates became less stable due to the smaller number of patients remaining at risk. These patterns support the use of survival analysis methods and highlight the clinical relevance of early treatment tolerance.

```{r figure2-all, fig.cap="Kaplan-Meier curves for time to dose-limiting toxicity by endpoint. Shaded bands represent 95\\% confidence intervals. Risk tables show the number of patients remaining under observation at each cycle.", fig.height=10, fig.width=6}
p_cap <- ggsurvplot(
  survfit(Surv(time, event) ~ 1, data = cap),
  data = cap, conf.int = TRUE, risk.table = TRUE,
  title = "CAP", xlab = "Chemotherapy cycles",
  ylab = "Probability toxicity-free",
  ggtheme = theme_minimal(base_size = 9), fontsize = 2.5,
  risk.table.height = 0.25
)
p_oxa <- ggsurvplot(
  survfit(Surv(time, event) ~ 1, data = oxa),
  data = oxa, conf.int = TRUE, risk.table = TRUE,
  title = "OXA", xlab = "Chemotherapy cycles",
  ylab = "Probability toxicity-free",
  ggtheme = theme_minimal(base_size = 9), fontsize = 2.5,
  risk.table.height = 0.25
)
p_capox <- ggsurvplot(
  survfit(Surv(time, event) ~ 1, data = capox),
  data = capox, conf.int = TRUE, risk.table = TRUE,
  title = "CAPOX", xlab = "Chemotherapy cycles",
  ylab = "Probability toxicity-free",
  ggtheme = theme_minimal(base_size = 9), fontsize = 2.5,
  risk.table.height = 0.25
)
arrange_ggsurvplots(list(p_cap, p_oxa, p_capox), ncol = 1, nrow = 3)
```

## Association between Liver Radiodensity and Toxicity

As shown in the descriptive analysis, both liver radiodensity and toxicity rates varied across hospitals, suggesting potential differences in measurement procedures, patient populations, or clinical practices. The association between liver radiodensity and time to toxicity was assessed using stratified Cox proportional hazards models, adjusted for age, sex, body mass index, smoking status, and treatment-specific starting dose. Stratification by hospital allows each hospital to have its own baseline hazard function, meaning that comparisons are made within hospitals rather than between hospitals. Table 4 presents the estimated hazard ratios for liver radiodensity across all three endpoints, and Table 5 presents the full model output including all covariates.

```{r table4}
hr_hu <- bind_rows(
  tidy(cox_cap,   exponentiate = TRUE, conf.int = TRUE) %>%
    filter(term == "liverHU_manual") %>% mutate(Endpoint = "CAP"),
  tidy(cox_oxa,   exponentiate = TRUE, conf.int = TRUE) %>%
    filter(term == "liverHU_manual") %>% mutate(Endpoint = "OXA"),
  tidy(cox_capox, exponentiate = TRUE, conf.int = TRUE) %>%
    filter(term == "liverHU_manual") %>% mutate(Endpoint = "CAPOX")
) %>%
  mutate(
    `Hazard Ratio (95% CI)` = paste0(
      round(estimate, 2), " (",
      round(conf.low, 2), "–",
      round(conf.high, 2), ")"
    ),
    `p-value` = round(p.value, 3),
    `Model` = "Stratified by hospital"
  ) %>%
  select(Endpoint, Model, `Hazard Ratio (95% CI)`, `p-value`)

kable(hr_hu, booktabs = TRUE,
      caption = "Adjusted association between liver radiodensity (per 1 HU) and time to toxicity") %>%
  kable_styling(latex_options = c("hold_position"), font_size = 10)
```

Hazard ratios for liver radiodensity were close to 1 across all three endpoints. For CAP, the estimated HR was 1.00 (95\% CI: 0.99 to 1.02, p = 0.376), indicating no association. For OXA, the HR was 1.01 (95\% CI: 1.00 to 1.02, p = 0.060), a borderline result that did not reach conventional statistical significance. For CAPOX, the HR was 1.01 (95\% CI: 1.00 to 1.02, p = 0.044), reaching statistical significance at the 5\% level. However, the effect size is very small: a one-unit increase in liver HU was associated with only a 1\% increase in the hazard of toxicity. This association is unlikely to be of clinical relevance.

```{r table5-full}
label_term <- function(x) {
  x %>%
    mutate(term = recode(term,
      "liverHU_manual" = "Liver HU",
      "age"            = "Age",
      "sex2"           = "Sex (male)",
      "BBMI"           = "BMI",
      "BSMOKER2"       = "Smoking (cat. 2)",
      "BSMOKER3"       = "Smoking (cat. 3)",
      "starting_dose"  = "Starting dose",
      "doseCAP"        = "Starting dose CAP",
      "doseOXA"        = "Starting dose OXA"
    ))
}

hr_full <- bind_rows(
  tidy(cox_cap,   exponentiate = TRUE, conf.int = TRUE) %>% label_term(),
  tidy(cox_oxa,   exponentiate = TRUE, conf.int = TRUE) %>% label_term(),
  tidy(cox_capox, exponentiate = TRUE, conf.int = TRUE) %>% label_term()
) %>%
  mutate(
    HR_CI = paste0(round(estimate, 2), " (", round(conf.low, 2), ", ", round(conf.high, 2), ")"),
    p     = ifelse(p.value < 0.001, "<0.001", as.character(round(p.value, 3)))
  ) %>%
  select(Covariate = term, `HR (95% CI)` = HR_CI, p)

kable(hr_full, booktabs = TRUE,
      caption = "Full adjusted hazard ratios from stratified Cox models (all covariates).",
      col.names = c("Covariate", "HR (95\\% CI)", "p-value")) %>%
  kable_styling(latex_options = c("hold_position"), font_size = 10) %>%
  pack_rows("CAP", 1, 7) %>%
  pack_rows("OXA", 8, 14) %>%
  pack_rows("CAPOX", 15, 22)
```

\noindent\small\textit{Note: All models stratified by hospital. Reference categories: sex = female; smoking = category 1.}
\normalsize

Although liver radiodensity was the primary exposure of interest, the full model output in Table 5 reveals that several covariates showed clinically plausible and statistically significant associations with toxicity risk, providing evidence that the models are detecting genuine signal. Age was significantly associated with toxicity in both the OXA model (HR = 1.03, 95\% CI: 1.01--1.05, p = 0.001) and the CAPOX model (HR = 1.03, 95\% CI: 1.01--1.05, p = 0.010), indicating that older patients had a moderately higher rate of toxicity per year of age. Sex showed a strong protective association in both the OXA model (HR = 0.48, 95\% CI: 0.34--0.68, p < 0.001) and the CAPOX model (HR = 0.63, 95\% CI: 0.45--0.89, p = 0.008), meaning that male patients experienced substantially lower toxicity rates than female patients after adjustment. Smoking status was significantly associated with higher toxicity risk in both OXA and CAPOX: in the OXA model, smoking category 2 had an HR of 1.98 (95\% CI: 1.11--3.56, p = 0.022) and category 3 had an HR of 2.18 (95\% CI: 1.21--3.90, p = 0.009); similar patterns were observed in the CAPOX model. Oxaliplatin starting dose was significantly associated with toxicity risk in the OXA model (HR = 1.01, 95\% CI: 1.01--1.02, p < 0.001). None of these associations were statistically significant in the CAP model. In contrast to the OXA and CAPOX endpoints, the CAP models showed generally weaker associations across all covariates, suggesting that capecitabine-related toxicity may be less driven by the patient characteristics included in this analysis.

## Model Diagnostics

The proportional hazards assumption was assessed using Schoenfeld residual tests for each of the three models. Table 6 reports the test statistic, degrees of freedom, and p-values for all variables in each model. A non-significant p-value (p > 0.05) indicates no evidence against the proportional hazards assumption. The GLOBAL row tests the assumption across all covariates simultaneously.

```{r table6-ph}
extract_ph <- function(ph, label) {
  df <- as.data.frame(ph$table)
  df$Variable <- rownames(df)
  df$Endpoint <- label
  df[, c("Endpoint", "Variable", "chisq", "df", "p")]
}

ph_combined <- bind_rows(
  extract_ph(ph_cap,   "CAP"),
  extract_ph(ph_oxa,   "OXA"),
  extract_ph(ph_capox, "CAPOX")
)

kable(ph_combined[, c("Variable", "chisq", "df", "p")],
      booktabs = TRUE, digits = 3, row.names = FALSE,
      caption = "Schoenfeld residual tests: proportional hazards assumption (all models).",
      col.names = c("Variable", "Chi-sq", "df", "p-value")) %>%
  kable_styling(latex_options = c("hold_position"), font_size = 10) %>%
  pack_rows("CAP",   1, nrow(as.data.frame(ph_cap$table))) %>%
  pack_rows("OXA",   nrow(as.data.frame(ph_cap$table)) + 1,
                     nrow(as.data.frame(ph_cap$table)) + nrow(as.data.frame(ph_oxa$table))) %>%
  pack_rows("CAPOX", nrow(as.data.frame(ph_cap$table)) + nrow(as.data.frame(ph_oxa$table)) + 1,
                     nrow(ph_combined))
```

All p-values are well above the conventional threshold of 0.05, indicating no evidence that the proportional hazards assumption is violated for any variable or for any of the three models. These findings support the validity of the Cox model and suggest that the estimated hazard ratios can be interpreted as constant over the treatment period.

# Summary of Findings

Taken together, the results show that while both liver radiodensity and toxicity rates vary across hospitals, liver radiodensity is not a strong or consistent predictor of chemotherapy-induced toxicity. After accounting for patient characteristics and hospital-level differences through stratification, hazard ratios for liver radiodensity were close to 1 across all endpoints. The association did not reach statistical significance for CAP (p = 0.376) or OXA (p = 0.060). For CAPOX, a statistically significant association was observed (HR = 1.01, p = 0.044); however, the effect size is very small and unlikely to be of clinical relevance. Other clinical factors, in particular sex, age, and smoking status, showed stronger and more consistent associations with toxicity risk across the OXA and CAPOX endpoints, suggesting that the models are sensitive to real clinical determinants of toxicity.

\newpage

# References

Iacovelli, R., Pietrantonio, F., Palazzo, A., Maggi, C., Ricchini, F., de Braud, F., et al. (2014). Incidence and relative risk of grade 3 and 4 diarrhoea in patients treated with capecitabine or 5-fluorouracil: A meta-analysis of published trials. *British Journal of Clinical Pharmacology, 78*(6), 1228--1237.

Kok, D. E., van Duijnhoven, F. J. B., Lubberman, F. J. E., McKay, J. A., van Lanen, A., Winkels, R. M., et al. (2024). Intake and biomarkers of folate and folic acid as determinants of chemotherapy-induced toxicities in patients with colorectal cancer: A cohort study. *The American Journal of Clinical Nutrition, 119*, 294--301.

Renting, L., Zwart, N. R. K., Ueland, P. M., McCann, A., Ulvik, A., van Halteren, H. K., Lubberman, F. J. E., Winkels, R. M., Kampman, E., & Kok, D. E. (2024). Vitamin B6 status and chronic chemotherapy-induced peripheral neuropathy: a prospective cohort study among patients with non-metastatic colorectal cancer receiving oxaliplatin-based chemotherapy. *BMJ Oncology, 3*: e000462.

Scheithauer, W., McKendrick, J., Begbie, S., Borner, M., Burns, W. I., Burris, H. A., et al. (2003). Oral capecitabine as an alternative to intravenous 5-fluorouracil-based adjuvant therapy for colon cancer: Safety results of a randomized phase III trial. *Annals of Oncology, 14*(12), 1735--1743.

Winkels, R. M., Heine-Bröring, R. C., van Zutphen, M., van Harten-Gerritsen, S., Kok, D. E., van Duijnhoven, F. J. B., et al. (2014). The COLON study: Colorectal cancer: Longitudinal, Observational study on Nutritional and lifestyle factors that may influence colorectal tumour recurrence, survival and quality of life. *BMC Cancer, 14*, 374.
