---
title: "Liver Radiodensity and Time to Dose-Limiting Toxicity"
author: "Bos Noah"
date: "`r Sys.Date()`"
output: pdf_document
editor_options: 
  markdown: 
    wrap: 72
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r, packages, message=FALSE, include = FALSE}
library('ipred')
library(TH.data) 
library(survival)
library(survminer)
library(hrbrthemes)
library(knitr)
library(forestmodel)
library(broom)
library(mice)
library(dplyr)
```

```{r data load}
load("data/Dataset_liverfat.RData")
df = dataset_leiden
```



# 1. introduction

The objective of this study is to investigate whether liver radiodensity (measured in Hounsfield Units, HU) is associated with time to dose-limiting toxicity (DLT) in patients receiving chemotherapy. Additionally, we assess whether hospital-level differences influence this association.

Time-to-event methods are used, where the outcome is the number of chemotherapy cycles completed before the occurrence of DLT.

# 2. Study Population and Endpoints
Three toxicity endpoints are considered:

- Capecitabine (CAP) toxicity
- Oxaliplatin (OXA) toxicity
- Combined CAPOX toxicity

Each endpoint defines a separate risk set, as not all patients received both drugs. Therefore, each endpoint requires a separate survival dataset to ensure the correct population at risk and time variable.

The time variable represents the number of chemotherapy cycles completed before DLT. A value of 0 indicates toxicity occurring during the first cycle.

# 3. Descriptive Statistics

## 3.1 Overall Event Rates

Across endpoints, the number of events and censored observations were as follows:

```{r}
endpoint_summary <- tibble(
  Endpoint = c("CAP", "OXA", "CAPOX"),
  Event_var = c("TIMT_CAP2", "TIMT_OXA2", "TIMT_CAPOX2")
) %>%
  rowwise() %>%
  mutate(
    N = sum(!is.na(df[[Event_var]])),
    Events = sum(df[[Event_var]] == 1, na.rm = TRUE),
    Censored = sum(df[[Event_var]] == 0, na.rm = TRUE)
  ) %>%
  ungroup() %>%
  dplyr::select(Endpoint, N, Events, Censored)

kable(endpoint_summary)
```
Event rates differed across endpoints, with OXA toxicity showing the highest proportion of events.

## 3.2 Distribution Across Hospitals

The number of patients per hospital ranged from 10 to 70. Event rates varied between hospitals (approximately 33% to 80%), suggesting heterogeneity in baseline toxicity risk.

```{r}
hospital_distribution = df %>%
  filter(!is.na(TIMT_CAP2)) %>%
  group_by(hospital) %>%
  summarise(
    N = n(),
    Events = sum(TIMT_CAP2 == 1),
    Event_pct = round(100 * Events / N, 1),
    Censored = sum(TIMT_CAP2 == 0)
  ) %>%
  arrange(desc(N))
kable(hospital_distribution)
```

These differences indicate that hospital-level variation may influence toxicity outcomes.

# 4. Liver Radiodensity Across Hospitals

## 4.1 Distribution of Liver HU

Liver radiodensity (HU) distributions across hospitals are shown below.

```{r}
library(ggplot2)

ggplot(df, aes(x = hospital, y = liverHU_manual)) +
  geom_boxplot() +
  labs(
    title = "Distribution of Liver radiodensity levels by Hospital",
    x = "Hospital",
    y = "Liver HU (manual)"
  ) +
  theme_minimal()
```

Visual inspection suggests differences in liver HU distributions between hospitals.

## 4.2 Formal Comparison

A one-way ANOVA was performed to assess differences in mean liver HU across hospitals.

```{r}
hospital_anova = anova(lm(liverHU_manual ~ hospital, data = df))
kable(hospital_anova)
```

Liver HU values differed significantly across hospitals (F = 12.1, p < 0.001), with approximately 26% of the variance explained by hospital.

## 4.3 Implications for Modeling

Because both liver HU and toxicity rates vary across hospitals, hospital may act as a confounder in the association between liver radiodensity and DLT. Therefore, hospital will be accounted for in the survival models using stratified Cox proportional hazards models, allowing hospital-specific baseline hazards.

# 5. Construction of Endpoint-Specific Survival Datasets

For each toxicity endpoint (CAP, OXA, CAPOX), a separate survival dataset was constructed to ensure the correct risk set and time definition.

The preprocessing steps include:

- Selection of the endpoint-specific event variable (TIMT_2) and time variable (cycle_till_).
- Restriction to patients with non-missing event and time (i.e., patients at risk for that endpoint).

Definition of:

- event = 1 if DLT occurred, 0 otherwise.
- time = number of chemotherapy cycles completed before DLT (0 = toxicity during first cycle).

Inclusion of appropriate dose covariates:

- CAP model → starting_doseCAP
- OXA model → starting_doseOXA
- CAPOX model → both CAP and OXA starting doses.

```{r}
make_endpoint <- function(data, endpoint = c("CAP", "OXA", "CAPOX")) {
  endpoint <- match.arg(endpoint)

  if (endpoint == "CAP") {
    event_var <- "TIMT_CAP2"
    time_var  <- "cycle_till_timt" #this is TIMT capecitabine 
  } else if (endpoint == "OXA") {
    event_var <- "TIMT_OXA2"
    time_var  <- "cycle_till_timtOXA"
  } else {
    event_var <- "TIMT_CAPOX2"
    time_var  <- "cycle_till_timtCAPOX"
  }

  out <- data %>%
    transmute(
      participant, hospital, age, sex, liverHU_manual, BBMI, BSMOKER,
      starting_doseCAP, starting_doseOXA,
      event = .data[[event_var]],
      time  = .data[[time_var]],
    ) %>%
    filter(!is.na(event), !is.na(time)) %>%
    mutate(
      event = as.integer(event == 1),
      time  = as.integer(time)
    )

  if (endpoint == "CAP") {
    out <- out %>%
      mutate(starting_dose = starting_doseCAP) %>%
      dplyr::select(-starting_doseCAP, -starting_doseOXA)
  } else if (endpoint == "OXA") {
    out <- out %>%
      dplyr::mutate(starting_dose = starting_doseOXA) %>%
      dplyr::select(-starting_doseCAP, -starting_doseOXA)
  } else {
    out <- out %>%
      dplyr::rename(doseCAP = starting_doseCAP, doseOXA = starting_doseOXA)
  }

  out
}

# Example
# df <- dataset_leiden
cap   <- make_endpoint(df, "CAP")
oxa   <- make_endpoint(df, "OXA")
capox <- make_endpoint(df, "CAPOX")
```

```{r}
structure_table <- tibble(
  Dataset = c("CAP", "OXA", "CAPOX"),
  Time_variable = c("cycle_till_timt",
                    "cycle_till_timtOXA",
                    "cycle_till_timtCAPOX"),
  Event_variable = c("TIMT_CAP2",
                     "TIMT_OXA2",
                     "TIMT_CAPOX2"),
  Dose_variables = c("starting_doseCAP",
                     "starting_doseOXA",
                     "starting_doseCAP + starting_doseOXA")
)

kable(structure_table, caption = "Structure of Endpoint-Specific Survival Datasets")
```


This ensures that each survival model uses the correct population, time scale, and treatment adjustment structure.

# 6. Statistical Analysis Plan

For each endpoint, time to first DLT will be analyzed using Cox proportional hazards models.

The primary exposure is liver radiodensity (per HU increase).

Models will adjust for:

- Age
- Sex
- BMI
- Smoking status
- Starting dose (endpoint-specific)

Hospital will be incorporated using stratified Cox regression to allow hospital-specific baseline hazards while estimating a common liver HU effect across hospitals.

Sensitivity analyses will include:

- Assessment of proportional hazards assumptions.
- Multiple imputation for missing covariates. (is this needed?)
- Evaluation of potential effect modification by hospital.
