---
title: "Liver Radiodensity and Time to Dose-Limiting Toxicity"
author: "Bos Noah"
date: "`r Sys.Date()`"
output: pdf_document
editor_options: 
  markdown: 
    wrap: 72
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r, packages, message=FALSE, include = FALSE}
library('ipred')
library(TH.data) 
library(survival)
library(survminer)
library(hrbrthemes)
library(knitr)
library(forestmodel)
library(broom)
library(mice)
library(dplyr)
```

```{r data load}
load("data/Dataset_liverfat.RData")
df = dataset_leiden
```



# 1. introduction

The objective of this study is to investigate whether liver radiodensity (measured in Hounsfield Units, HU) is associated with time to dose-limiting toxicity (DLT) in patients receiving chemotherapy. Additionally, we assess whether hospital-level differences influence this association.

Time-to-event methods are used, where the outcome is the number of chemotherapy cycles completed before the occurrence of DLT.

# 2. Study Population and Endpoints
Three toxicity endpoints are considered:

- Capecitabine (CAP) toxicity
- Oxaliplatin (OXA) toxicity
- Combined CAPOX toxicity

Each endpoint defines a separate risk set, as not all patients received both drugs. Therefore, each endpoint requires a separate survival dataset to ensure the correct population at risk and time variable.

The time variable represents the number of chemotherapy cycles completed before DLT. A value of 0 indicates toxicity occurring during the first cycle.

# 3. Descriptive Statistics

## 3.1 Overall Event Rates

Across endpoints, the number of events and censored observations were as follows:

```{r}
endpoint_summary <- tibble(
  Endpoint = c("CAP", "OXA", "CAPOX"),
  Event_var = c("TIMT_CAP2", "TIMT_OXA2", "TIMT_CAPOX2")
) %>%
  rowwise() %>%
  mutate(
    N = sum(!is.na(df[[Event_var]])),
    Events = sum(df[[Event_var]] == 1, na.rm = TRUE),
    Censored = sum(df[[Event_var]] == 0, na.rm = TRUE)
  ) %>%
  ungroup() %>%
  dplyr::select(Endpoint, N, Events, Censored)

kable(endpoint_summary)
```
Event rates differed across endpoints, with OXA showing the highest proportion of toxicity events. This indicates heterogeneity in toxicity risk depending on treatment.

## 3.2 Distribution Across Hospitals

The number of patients per hospital ranged from 10 to 70. Event rates varied between hospitals (approximately 33% to 80%), suggesting heterogeneity in baseline toxicity risk.

```{r}
hospital_distribution = df %>%
  filter(!is.na(TIMT_CAP2)) %>%
  group_by(hospital) %>%
  summarise(
    N = n(),
    Events = sum(TIMT_CAP2 == 1),
    Event_pct = round(100 * Events / N, 1),
    Censored = sum(TIMT_CAP2 == 0)
  ) %>%
  arrange(desc(N))
kable(hospital_distribution)
```

Event rates varied across hospitals, suggesting differences in patient populations or treatment practices. This justifies adjusting for hospital in the survival models.

# 4. Liver Radiodensity Across Hospitals

## 4.1 Distribution of Liver HU

Liver radiodensity (HU) distributions across hospitals are shown below.

```{r}
library(ggplot2)

ggplot(df, aes(x = hospital, y = liverHU_manual)) +
  geom_boxplot() +
  labs(
    title = "Distribution of Liver radiodensity levels by Hospital",
    x = "Hospital",
    y = "Liver HU (manual)"
  ) +
  theme_minimal()
```

Liver radiodensity distributions differed across hospitals, indicating potential confounding. Hospital adjustment was therefore included in the Cox models.

## 4.2 Formal Comparison

A one-way ANOVA was performed to assess differences in mean liver HU across hospitals.

```{r}
hospital_anova = anova(lm(liverHU_manual ~ hospital, data = df))
kable(hospital_anova)
```

Liver HU values differed significantly across hospitals (F = 12.1, p < 0.001), with approximately 26% of the variance explained by hospital.

Mean liver radiodensity differed significantly across hospitals, supporting the need to account for hospital-level variation in the analysis.

## 4.3 Implications for Modeling

Because both liver HU and toxicity rates vary across hospitals, hospital may act as a confounder in the association between liver radiodensity and DLT. Therefore, hospital will be accounted for in the survival models using stratified Cox proportional hazards models, allowing hospital-specific baseline hazards.

# 5. Construction of Endpoint-Specific Survival Datasets

For each toxicity endpoint (CAP, OXA, CAPOX), a separate survival dataset was constructed to ensure the correct risk set and time definition.

The preprocessing steps include:

- Selection of the endpoint-specific event variable (TIMT_2) and time variable (cycle_till_).
- Restriction to patients with non-missing event and time (i.e., patients at risk for that endpoint).

Definition of:

- event = 1 if DLT occurred, 0 otherwise.
- time = number of chemotherapy cycles completed before DLT (0 = toxicity during first cycle).

Inclusion of appropriate dose covariates:

- CAP model → starting_doseCAP
- OXA model → starting_doseOXA
- CAPOX model → both CAP and OXA starting doses.

```{r}
make_endpoint <- function(data, endpoint = c("CAP", "OXA", "CAPOX")) {
  endpoint <- match.arg(endpoint)

  if (endpoint == "CAP") {
    event_var <- "TIMT_CAP2"
    time_var  <- "cycle_till_timt" #this is TIMT capecitabine 
  } else if (endpoint == "OXA") {
    event_var <- "TIMT_OXA2"
    time_var  <- "cycle_till_timtOXA"
  } else {
    event_var <- "TIMT_CAPOX2"
    time_var  <- "cycle_till_timtCAPOX"
  }

  out <- data %>%
    transmute(
      participant, hospital, age, sex, liverHU_manual, BBMI, BSMOKER,
      starting_doseCAP, starting_doseOXA,
      event = .data[[event_var]],
      time  = .data[[time_var]],
    ) %>%
    filter(!is.na(event), !is.na(time)) %>%
    mutate(
      event = as.integer(event == 1),
      time  = as.integer(time)
    )

  if (endpoint == "CAP") {
    out <- out %>%
      mutate(starting_dose = starting_doseCAP) %>%
      dplyr::select(-starting_doseCAP, -starting_doseOXA)
  } else if (endpoint == "OXA") {
    out <- out %>%
      dplyr::mutate(starting_dose = starting_doseOXA) %>%
      dplyr::select(-starting_doseCAP, -starting_doseOXA)
  } else {
    out <- out %>%
      dplyr::rename(doseCAP = starting_doseCAP, doseOXA = starting_doseOXA)
  }

  out
}

# Example
# df <- dataset_leiden
cap   <- make_endpoint(df, "CAP")
oxa   <- make_endpoint(df, "OXA")
capox <- make_endpoint(df, "CAPOX")

cap$liverHU_manual   <- as.numeric(cap$liverHU_manual)
oxa$liverHU_manual   <- as.numeric(oxa$liverHU_manual)
capox$liverHU_manual <- as.numeric(capox$liverHU_manual)
```

```{r}
structure_table <- tibble(
  Dataset = c("CAP", "OXA", "CAPOX"),
  Time_variable = c("cycle_till_timt",
                    "cycle_till_timtOXA",
                    "cycle_till_timtCAPOX"),
  Event_variable = c("TIMT_CAP2",
                     "TIMT_OXA2",
                     "TIMT_CAPOX2"),
  Dose_variables = c("starting_doseCAP",
                     "starting_doseOXA",
                     "starting_doseCAP + starting_doseOXA")
)

kable(structure_table, caption = "Structure of Endpoint-Specific Survival Datasets")
```

Separate survival datasets were constructed for each endpoint to ensure correct risk sets and treatment adjustment.

# 6. Kaplan-Meier curves


```{r}
plot_km <- function(data, title){

  fit <- survfit(Surv(time, event) ~ 1, data = data)

  ggsurvplot(
    fit,
    data = data,  
    conf.int = TRUE,
    risk.table = TRUE,
    ggtheme = theme_minimal(),
    title = title,
    xlab = "Chemotherapy cycles",
    ylab = "Probability no DLT"
  )
}

plot_km(cap, "CAP toxicity")
plot_km(oxa, "OXA toxicity")
plot_km(capox, "CAPOX toxicity")
```
Kaplan–Meier curves show the unadjusted probability of remaining free from dose-limiting toxicity over chemotherapy cycles. For all endpoints, toxicity occurs primarily during the first treatment cycles. This indicates that time-to-event analysis is appropriate and that early treatment tolerance is clinically relevant.

# 7. Cox proportional hazards models

```{r}
cox_cap <- coxph(
  Surv(time, event) ~ liverHU_manual + age + sex + BBMI +
    BSMOKER + starting_dose + strata(hospital),
  data = cap
)

cox_oxa <- coxph(
  Surv(time, event) ~ liverHU_manual + age + sex + BBMI +
    BSMOKER + starting_dose + strata(hospital),
  data = oxa
)

cox_capox <- coxph(
  Surv(time, event) ~ liverHU_manual + age + sex + BBMI +
    BSMOKER + doseCAP + doseOXA + strata(hospital),
  data = capox
)

summary(cox_cap)
summary(cox_oxa)
summary(cox_capox)
```
Cox proportional hazards models were fitted to evaluate whether liver radiodensity is associated with time to dose-limiting toxicity while adjusting for patient characteristics and starting dose. Hospital was included as a stratification factor to account for center-level differences. This approach estimates the independent effect of liver radiodensity.



# 8. Hazard ratio table

```{r}

hr_table <- bind_rows(
  tidy(cox_cap, exponentiate = TRUE, conf.int = TRUE) %>% mutate(Model="CAP"),
  tidy(cox_oxa, exponentiate = TRUE, conf.int = TRUE) %>% mutate(Model="OXA"),
  tidy(cox_capox, exponentiate = TRUE, conf.int = TRUE) %>% mutate(Model="CAPOX")
)

kable(hr_table, digits = 3)
```

The hazard ratio table summarizes adjusted associations. Hazard ratios above 1 indicate higher toxicity risk, while values below 1 indicate lower risk. Liver radiodensity shows hazard ratios close to 1 across endpoints, suggesting a limited independent association with toxicity after adjustment.

# 9. Forest plots

```{r}

forest_model(cox_cap)
forest_model(cox_oxa)
forest_model(cox_capox)
```
Forest plots visualize the magnitude and precision of adjusted effects. Clinical covariates such as dose and patient characteristics show stronger associations with toxicity than liver radiodensity. This suggests that liver radiodensity is not a dominant predictor of toxicity risk.

# 10. PH assumption
```{r}
ph_cap <- cox.zph(cox_cap)
ph_oxa <- cox.zph(cox_oxa)
ph_capox <- cox.zph(cox_capox)

kable(as.data.frame(ph_cap$table))
kable(as.data.frame(ph_oxa$table))
kable(as.data.frame(ph_capox$table))
```
The proportional hazards assumption was evaluated using Schoenfeld residual tests. Non-significant results indicate no strong evidence that hazard ratios change over time. This supports the validity of the Cox model estimates.

```{r}
par(mfrow=c(3,3))
plot(ph_cap)
plot(ph_oxa)
plot(ph_capox)
par(mfrow=c(1,1))
```

Schoenfeld residual plots show the stability of covariate effects over time. No clear systematic trends were observed, suggesting that proportional hazards assumptions are reasonably met.


# 11 conclusion

This study evaluated whether liver radiodensity predicts time to dose-limiting toxicity. Kaplan–Meier curves showed that most toxicity occurs early in treatment. After adjustment for patient characteristics, dose, and hospital, liver radiodensity demonstrated only a small association with toxicity risk. Clinical factors and treatment dose showed stronger effects. These findings suggest that liver radiodensity alone is unlikely to be a clinically useful predictor of chemotherapy toxicity.

## Clinical interpretation of effect size

Question for Deborah:

To aid clinical interpretation, the effect of liver radiodensity can be considered over a 10-HU difference. For example, a hazard ratio of 0.99 per 1 HU corresponds to approximately 0.90 per 10 HU, indicating a modest reduction in toxicity risk. Similarly, a hazard ratio of 1.01 per 1 HU corresponds to approximately 1.10 per 10 HU, indicating a modest increase. These effect sizes suggest limited clinical impact of liver radiodensity.

