---
title: "Liver Radiodensity and Time to Dose-Limiting Toxicity"
author: "Bos Noah"
date: "`r Sys.Date()`"
output:
  beamer_presentation:
    theme: "Madrid"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.width = 6,
  fig.height = 4
)

library(survival)
library(survminer)
library(dplyr)
library(knitr)
library(broom)
library(ggplot2)
library(forestmodel)
```

```{r}
load("data/Dataset_liverfat.RData")
df = dataset_leiden
```

# Objective

- Evaluate association between liver HU and toxicity
- Time-to-event: cycles until dose-limiting toxicity
- Adjust for clinical factors and hospital

**Conclusion:** Study assesses predictive value of liver radiodensity.

# Endpoints

- CAP toxicity
- OXA toxicity
- CAPOX toxicity

**Conclusion:** Three independent survival analyses performed.

# Event rates

```{r}
endpoint_summary <- tibble(
  Endpoint = c("CAP", "OXA", "CAPOX"),
  Event_var = c("TIMT_CAP2", "TIMT_OXA2", "TIMT_CAPOX2")
) %>%
  rowwise() %>%
  mutate(
    N = sum(!is.na(df[[Event_var]])),
    Events = sum(df[[Event_var]] == 1, na.rm = TRUE),
    Censored = sum(df[[Event_var]] == 0, na.rm = TRUE)
  ) %>%
  ungroup() %>%
  select(Endpoint, N, Events, Censored)

kable(endpoint_summary)
```

**Conclusion:** Toxicity frequency differs across treatments.

# Liver HU by hospital

```{r}
ggplot(df, aes(hospital, liverHU_manual)) +
  geom_boxplot() +
  theme_minimal()
```


```{r}
make_endpoint <- function(data, endpoint){

  if(endpoint=="CAP"){
    event <- data$TIMT_CAP2
    time  <- data$cycle_till_timt
    dose  <- data$starting_doseCAP
  }

  if(endpoint=="OXA"){
    event <- data$TIMT_OXA2
    time  <- data$cycle_till_timtOXA
    dose  <- data$starting_doseOXA
  }

  if(endpoint=="CAPOX"){
    event <- data$TIMT_CAPOX2
    time  <- data$cycle_till_timtCAPOX
  }

  out <- data.frame(
    time,
    event,
    liverHU_manual=data$liverHU_manual,
    age=data$age,
    sex=data$sex,
    BBMI=data$BBMI,
    BSMOKER=data$BSMOKER,
    hospital=data$hospital
  )

  if(endpoint!="CAPOX"){
    out$starting_dose <- dose
  }

  out <- out[complete.cases(out),]
  out
}

cap <- make_endpoint(df,"CAP")
oxa <- make_endpoint(df,"OXA")
capox <- make_endpoint(df,"CAPOX")
```


# Kaplan-Meier CAP

```{r}
ggsurvplot(
survfit(Surv(time,event)~1,data=cap),
risk.table=T
)
```


# Kaplan-Meier OXA

```{r}
ggsurvplot(
survfit(Surv(time,event)~1,data=oxa),
risk.table=T
)
```
The probability of remaining toxicity-free decreases steadily. However, estimates after 10–12 cycles are unstable due to very few patients remaining at risk.

# Kaplan-Meier CAPOX

```{r}
ggsurvplot(
survfit(Surv(time,event)~1,data=capox),
risk.table=T
)
```
CAPOX shows a similar pattern, with most toxicities occurring during early treatment cycles. Late-cycle estimates are based on very small numbers and should be interpreted cautiously.

# Cox models

```{r}
cox_cap <- coxph(
Surv(time,event)~liverHU_manual+age+sex+BBMI+
BSMOKER+starting_dose+strata(hospital),
data=cap)

cox_oxa <- coxph(
Surv(time,event)~liverHU_manual+age+sex+BBMI+
BSMOKER+starting_dose+strata(hospital),
data=oxa)

cox_capox <- coxph(
Surv(time,event)~liverHU_manual+age+sex+BBMI+
BSMOKER+strata(hospital),
data=capox)
```

**Conclusion KM Curves**

Across all treatments, most dose-limiting toxicities occur during early chemotherapy cycles, with decreasing numbers of patients at risk in later cycles leading to increased uncertainty.

Few patients remain under observation at later cycles, leading to wider uncertainty.

This justifies survival analysis

**Next step**

Multivariable Cox models fitted for all three endpoints.

Hospital stratification allows baseline toxicity risk to differ between centers without estimating separate coefficients.

- each hospital has its own baseline hazard
- but same covariate effects

# Hazard ratio table

```{r }
hr_table <- bind_rows(
  tidy(cox_cap, exponentiate = TRUE, conf.int = TRUE) %>% mutate(Model="CAP"),
  tidy(cox_oxa, exponentiate = TRUE, conf.int = TRUE) %>% mutate(Model="OXA"),
  tidy(cox_capox, exponentiate = TRUE, conf.int = TRUE) %>% mutate(Model="CAPOX")
)

kable(hr_table, digits = 2) %>%
  kableExtra::kable_styling(font_size = 7)
```

**Conclusion:** Liver HU effect close to null across endpoints.

# Forest plot CAP

```{r}
forest_model(cox_cap)
```

**Conclusion:** CAP model shows small liver HU effect.

# Forest plot OXA

```{r}
forest_model(cox_oxa)
```

**Conclusion:** OXA model confirms weak association.

# Forest plot CAPOX

```{r}
forest_model(cox_capox)
```

**Conclusion:** CAPOX model also shows minimal effect.

# PH assumption

The proportional hazards assumption was evaluated using Schoenfeld residual tests. Non-significant results indicate no strong evidence that hazard ratios change over time. This supports the validity of the Cox model estimates.

```{r}
ph_capox <- cox.zph(cox_capox)
kable(as.data.frame(ph_capox$table))
```

**Conclusion:** No major PH violations.

# Overall conclusion

- Liver radiodensity weakly associated with toxicity
- Effects small across all endpoints
- Dose and clinical variables stronger predictors
- Hospital differences accounted for

**Final conclusion:** Liver radiodensity is not a strong predictor of chemotherapy toxicity.
