---
title: "Does Liver Radiodensity Predict Chemotherapy Toxicity?"
subtitle: "Time-to-event analysis across CAP, OXA, and CAPOX"
author: "Noah Bos & Abdullah Alam"
date: "`r Sys.Date()`"

output:
  beamer_presentation:
    theme: "Madrid"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 5
)

library(survival)
library(survminer)
library(dplyr)
library(knitr)
library(broom)
library(ggplot2)
library(forestmodel)


```

```{r}
load("data/Dataset_liverfat.RData")
df = dataset_leiden
```

# Study Objective

- Assess whether liver radiodensity (HU) predicts time to dose-limiting toxicity  
- Outcome: number of cycles until toxicity occurs  
- Adjusted for patient characteristics and hospital differences  

**Key takeaway:** Liver HU is evaluated for its added predictive value in clinical decision-making.

# Toxicity Endpoints Analyzed Separately

- CAP toxicity  
- OXA toxicity  
- CAPOX toxicity  

**Key takeaway:** Each treatment is evaluated using a separate time-to-event model.

# Observed Toxicity Rates Differ by Treatment

```{r}
endpoint_summary <- tibble(
  Endpoint = c("CAP", "OXA", "CAPOX"),
  Event_var = c("TIMT_CAP2", "TIMT_OXA2", "TIMT_CAPOX2")
) %>%
  rowwise() %>%
  mutate(
    N = sum(!is.na(df[[Event_var]])),
    Events = sum(df[[Event_var]] == 1, na.rm = TRUE),
    Censored = sum(df[[Event_var]] == 0, na.rm = TRUE)
  ) %>%
  ungroup() %>%
  select(Endpoint, N, Events, Censored)

kable(endpoint_summary)
```

**Key takeaway:** Event rates differ across treatments, with sufficient events to support survival analysis.

# Liver Radiodensity Varies Across Hospitals

```{r}
ggplot(df, aes(hospital, liverHU_manual)) +
  geom_boxplot() +
  theme_minimal()
```

**Key takeaway:** Liver HU distributions differ between hospitals, suggesting baseline differences that are accounted for through stratification.


```{r}
make_endpoint <- function(data, endpoint){

  if(endpoint=="CAP"){
    event <- data$TIMT_CAP2
    time  <- data$cycle_till_timt
    dose  <- data$starting_doseCAP
  }

  if(endpoint=="OXA"){
    event <- data$TIMT_OXA2
    time  <- data$cycle_till_timtOXA
    dose  <- data$starting_doseOXA
  }

  if(endpoint=="CAPOX"){
    event <- data$TIMT_CAPOX2
    time  <- data$cycle_till_timtCAPOX
  }

  out <- data.frame(
    time,
    event,
    liverHU_manual=data$liverHU_manual,
    age=data$age,
    sex=data$sex,
    BBMI=data$BBMI,
    BSMOKER=data$BSMOKER,
    hospital=data$hospital
  )

  if(endpoint!="CAPOX"){
    out$starting_dose <- dose
  }

  out <- out[complete.cases(out),]
  out
}

cap <- make_endpoint(df,"CAP")
oxa <- make_endpoint(df,"OXA")
capox <- make_endpoint(df,"CAPOX")
```


# Time to Toxicity: CAP

```{r}
ggsurvplot(
survfit(Surv(time,event)~1,data=cap),
risk.table=T
)
```
**Key takeaway:** The probability of remaining toxicity-free declines over cycles, with most events occurring early in treatment.


# Time to Toxicity: OXA

```{r}
ggsurvplot(
survfit(Surv(time,event)~1,data=oxa),
risk.table=T
)
```
**Key takeaway:** Toxicity risk is highest in early cycles. Estimates in later cycles are less reliable due to few patients remaining at risk.


# Time to Toxicity: CAPOX

```{r}
ggsurvplot(
survfit(Surv(time,event)~1,data=capox),
risk.table=T
)
```
**Key takeaway:** A similar pattern is observed, with most toxicities occurring early and increasing uncertainty in later cycles.

# Why Use Cox Models?

```{r}
cox_cap <- coxph(
Surv(time,event)~liverHU_manual+age+sex+BBMI+
BSMOKER+starting_dose+strata(hospital),
data=cap)

cox_oxa <- coxph(
Surv(time,event)~liverHU_manual+age+sex+BBMI+
BSMOKER+starting_dose+strata(hospital),
data=oxa)

cox_capox <- coxph(
Surv(time,event)~liverHU_manual+age+sex+BBMI+
BSMOKER+strata(hospital),
data=capox)
```

- Kaplan-Meier curves describe when toxicity occurs  
- They do not quantify effects of predictors  

- Cox models estimate the association between liver HU and toxicity  
- Stratified by hospital to allow different baseline risks  

**Key takeaway:** Cox models quantify whether liver radiodensity predicts toxicity after adjustment.

# Association Between Liver HU and Toxicity

```{r }
hr_table <- bind_rows(
  tidy(cox_cap, exponentiate = TRUE, conf.int = TRUE) %>% mutate(Model="CAP"),
  tidy(cox_oxa, exponentiate = TRUE, conf.int = TRUE) %>% mutate(Model="OXA"),
  tidy(cox_capox, exponentiate = TRUE, conf.int = TRUE) %>% mutate(Model="CAPOX")
)

kable(hr_table, digits = 2) %>%
  kableExtra::kable_styling(font_size = 7)
```

**Key takeaway:** Hazard ratios for liver HU are close to 1 across all endpoints, with no statistically significant associations observed.

# CAP: No Meaningful Association with Liver HU

```{r}
forest_model(cox_cap)
```

**Key takeaway:** The estimated effect of liver HU is close to null with no statistically significant association.

# OXA: Small Trend, Not Statistically significant

```{r}
forest_model(cox_oxa)
```

**Key takeaway:** A slight association is observed, but the effect size is small and not significant


# CAPOX: No Consistent Association Observed

```{r}
forest_model(cox_capox)
```

**Key takeaway:** Estimates are close to null, with no clear evidence of an association.

# Model Assumptions Hold

- No evidence of violation of the proportional hazards assumption  
- Hazard ratios appear stable over time  

```{r}
ph_capox <- cox.zph(cox_capox)
kable(as.data.frame(ph_capox$table))
```

**Key takeaway:** The Cox model assumptions are satisfied, supporting the validity of the results.

# Overall Interpretation

- Liver radiodensity shows no consistent or clinically meaningful association with time to toxicity  
- Effect sizes are small and not statistically significant across endpoints  
- Other clinical variables (e.g., age, smoking, dose) show stronger associations  

**Key takeaway:** Liver radiodensity does not appear to add meaningful predictive value beyond standard clinical variables.

# Future study

No clear association was observed when liver radiodensity was modeled continuously across its full range. A possible follow-up analysis would be to assess whether toxicity risk differs only at clinically meaningful thresholds or extreme radiodensity ranges.
