---
title: "Liver Radiodensity and Time to Dose-Limiting Toxicity"
author: "Bos Noah"
date: "`r Sys.Date()`"
output:
  beamer_presentation:
    theme: Madrid
fontsize: 11pt
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center"
)

knitr::opts_chunk$set(
  fig.width = 6,
  fig.height = 4,
  out.width = "90%"
)
```

```{r packages, include=FALSE}
library(ipred)
library(TH.data)
library(survival)
library(survminer)
library(hrbrthemes)
library(knitr)
library(forestmodel)
library(broom)
library(mice)
library(dplyr)
library(ggplot2)
```

```{r data-load}
load("data/Dataset_liverfat.RData")
df <- dataset_leiden
```

# Study Objective

## Research Question

Goal: investigate whether **liver radiodensity (HU)** is associated with  
**time to dose-limiting toxicity (DLT)** during chemotherapy.

Secondary objective: evaluate whether **hospital-level variation** affects this relationship.

Method: **time-to-event analysis using Cox proportional hazards models.**

---

# Study Endpoints

Three toxicity endpoints:

- CAP toxicity
- OXA toxicity
- CAPOX toxicity

Not all patients received both drugs → **different risk sets**.

Time variable: **number of chemotherapy cycles until DLT**

- `0` = toxicity during the first cycle.

---

# Descriptive Statistics

## Overall Event Rates

```{r}
endpoint_summary <- tibble(
  Endpoint = c("CAP", "OXA", "CAPOX"),
  Event_var = c("TIMT_CAP2", "TIMT_OXA2", "TIMT_CAPOX2")
) %>%
  rowwise() %>%
  mutate(
    N = sum(!is.na(df[[Event_var]])),
    Events = sum(df[[Event_var]] == 1, na.rm = TRUE),
    Censored = sum(df[[Event_var]] == 0, na.rm = TRUE)
  ) %>%
  ungroup() %>%
  select(Endpoint, N, Events, Censored)

kable(endpoint_summary)
```

Event rates differ across endpoints.

---

# Hospital Distribution

## Patients per Hospital

```{r}
hospital_distribution <- df %>%
  filter(!is.na(TIMT_CAP2)) %>%
  group_by(hospital) %>%
  summarise(
    N = n(),
    Events = sum(TIMT_CAP2 == 1),
    Event_pct = round(100 * Events / N, 1),
    Censored = sum(TIMT_CAP2 == 0)
  ) %>%
  arrange(desc(N))

kable(hospital_distribution)
```

Hospitals range roughly **10–70 patients** with varying event rates.

---

# Liver Radiodensity

## HU Distribution Across Hospitals

```{r fig.width=4, fig.height=2}
ggplot(df, aes(x = hospital, y = liverHU_manual)) +
  geom_boxplot() +
  labs(
    title = "Distribution of Liver Radiodensity by Hospital",
    x = "Hospital",
    y = "Liver HU"
  ) +
  theme_minimal()
```

Visual inspection suggests differences between hospitals.

---

# ANOVA Test

```{r}
hospital_anova <- anova(lm(liverHU_manual ~ hospital, data = df))
kable(hospital_anova)
```

Interpretation:

Significant differences in liver HU across hospitals.

---

# Modeling Implication

Both

- **Liver HU**
- **DLT rates**

vary by hospital → potential **confounding**.

Solution:

Use **stratified Cox proportional hazards models**

This allows:

- hospital-specific baseline hazards
- common liver HU effect.

---

# Survival Dataset Construction

Each endpoint uses a **separate survival dataset**.

Definitions:

- `event = 1` → DLT occurred
- `event = 0` → censored
- `time` → cycles until DLT

---


```{r}
make_endpoint <- function(data, endpoint = c("CAP", "OXA", "CAPOX")) {

  endpoint <- match.arg(endpoint)

  if (endpoint == "CAP") {
    event_var <- "TIMT_CAP2"
    time_var  <- "cycle_till_timt"
  } else if (endpoint == "OXA") {
    event_var <- "TIMT_OXA2"
    time_var  <- "cycle_till_timtOXA"
  } else {
    event_var <- "TIMT_CAPOX2"
    time_var  <- "cycle_till_timtCAPOX"
  }

  out <- data %>%
    transmute(
      participant,
      hospital,
      age,
      sex,
      liverHU_manual,
      BBMI,
      BSMOKER,
      starting_doseCAP,
      starting_doseOXA,
      event = .data[[event_var]],
      time  = .data[[time_var]]
    ) %>%
    filter(!is.na(event), !is.na(time)) %>%
    mutate(
      event = as.integer(event == 1),
      time = as.integer(time)
    )

  if (endpoint == "CAP") {
    out <- out %>%
      mutate(starting_dose = starting_doseCAP) %>%
      select(-starting_doseCAP, -starting_doseOXA)

  } else if (endpoint == "OXA") {
    out <- out %>%
      mutate(starting_dose = starting_doseOXA) %>%
      select(-starting_doseCAP, -starting_doseOXA)

  } else {
    out <- out %>%
      rename(doseCAP = starting_doseCAP,
             doseOXA = starting_doseOXA)
  }

  out
}

cap   <- make_endpoint(df, "CAP")
oxa   <- make_endpoint(df, "OXA")
capox <- make_endpoint(df, "CAPOX")
```


# Endpoint Structure

```{r, size=4}
structure_table <- tibble(
  Dataset = c("CAP", "OXA", "CAPOX"),
  Time_variable = c("cycle_till_timt",
                    "cycle_till_timtOXA",
                    "cycle_till_timtCAPOX"),
  Event_variable = c("TIMT_CAP2",
                     "TIMT_OXA2",
                     "TIMT_CAPOX2"),
  Dose_variables = c("starting_doseCAP",
                     "starting_doseOXA",
                     "starting CAP + OXA")
)

kable(structure_table)
```

---

# 7. Kaplan–Meier Curves

Kaplan–Meier curves were constructed to visualise the overall time-to-toxicity distribution.

```{r}
km_cap <- survfit(Surv(time, event) ~ 1, data = cap)
km_oxa <- survfit(Surv(time, event) ~ 1, data = oxa)
km_capox <- survfit(Surv(time, event) ~ 1, data = capox)

ggsurvplot(km_cap, risk.table = TRUE, title = "CAP toxicity")
ggsurvplot(km_oxa, risk.table = TRUE, title = "OXA toxicity")
ggsurvplot(km_capox, risk.table = TRUE, title = "CAPOX toxicity")

```
# 8. Cox proportional hazards models


Multivariable Cox proportional hazards models were fitted for each endpoint, adjusting for:
	•	Liver radiodensity (HU)
	•	Age
	•	Sex
	•	BMI
	•	Smoking
	•	Starting dose
	•	Hospital (stratified baseline hazards)

```{r}
cox_cap <- coxph(
  Surv(time, event) ~ liverHU_manual + age + sex + BBMI +
    BSMOKER + starting_dose + strata(hospital),
  data = cap
)

cox_oxa <- coxph(
  Surv(time, event) ~ liverHU_manual + age + sex + BBMI +
    BSMOKER + starting_dose + strata(hospital),
  data = oxa
)

cox_capox <- coxph(
  Surv(time, event) ~ liverHU_manual + age + sex + BBMI +
    BSMOKER + doseCAP + doseOXA + strata(hospital),
  data = capox
)

summary(cox_cap)
summary(cox_oxa)
summary(cox_capox)
```

# 9. Hazard ratio table

```{r}
hr_table <- bind_rows(
  tidy(cox_cap, exponentiate = TRUE, conf.int = TRUE) %>% mutate(Model="CAP"),
  tidy(cox_oxa, exponentiate = TRUE, conf.int = TRUE) %>% mutate(Model="OXA"),
  tidy(cox_capox, exponentiate = TRUE, conf.int = TRUE) %>% mutate(Model="CAPOX")
)

kable(hr_table, digits = 3)
```

# 10. Interpretation per 10 HU
To improve clinical interpretability, hazard ratios were converted to a 10 HU difference.

```{r}
hr_10HU <- hr_table %>%
  filter(term == "liverHU_manual") %>%
  mutate(
    HR_10HU = estimate^10,
    conf.low_10HU = conf.low^10,
    conf.high_10HU = conf.high^10
  )

kable(hr_10HU)
```
Even for a 10 HU difference, the effect size remained small, suggesting limited clinical relevance.

# 11. forest plots
```{r}
forest_model(cox_cap)
forest_model(cox_oxa)
forest_model(cox_capox)
```
Forest plots illustrate that liver HU has a smaller effect compared to clinical covariates such as dose.

# 12. Proportional Hazards assumption

```{r}
ph_cap <- cox.zph(cox_cap)
ph_oxa <- cox.zph(cox_oxa)
ph_capox <- cox.zph(cox_capox)

kable(as.data.frame(ph_cap$table))
kable(as.data.frame(ph_oxa$table))
kable(as.data.frame(ph_capox$table))
```

```{r}
plot(ph_cap)
plot(ph_oxa)
plot(ph_capox)
```
No major violations of the proportional hazards assumption were observed.

# 13. Adjusted survival curves

Adjusted survival curves were constructed to visualise the effect of liver radiodensity while holding covariates constant.
```{r}
cap$liver_group <- ifelse(cap$liverHU_manual < median(cap$liverHU_manual), "Low HU", "High HU")

fit_adj <- survfit(
  coxph(
    Surv(time, event) ~ liver_group + age + sex + BBMI +
      BSMOKER + starting_dose + strata(hospital),
    data = cap
  )
)

ggsurvplot(fit_adj, data = cap, risk.table = TRUE)
```

# 14. Summary of findings
- Liver radiodensity showed only a weak association with toxicity risk
- Hazard ratios were close to 1 across all endpoints
- Even a 10 HU difference resulted in minimal change in risk
- Starting dose and clinical variables had larger effects
- Hospital differences were accounted for using stratification

# 15. conclusion

Liver radiodensity is not a strong independent predictor of dose-limiting chemotherapy toxicity.
Clinical factors and treatment dose appear more important determinants of toxicity risk.
