---
title: "Liver Radiodensity and Chemotherapy Toxicity"
subtitle: "Hospital-Level Variation in a Multi-centre Cohort"
author: "Noah Bos \\& Abdullah Alam"
institute: "COLON Study · Statistical Consulting Project"
date: "`r Sys.Date()`"
knit: (function(input, ...) rmarkdown::render(input, output_dir = here::here("reports"), output_file = "slides.pdf"))
classoption: "aspectratio=169"
output:
  beamer_presentation:
    theme: "metropolis"
    keep_tex: false
    slide_level: 2
header-includes:
  - \usepackage{booktabs}
  - \usepackage{amsmath}
  - \usepackage{tikz}
  - \usetikzlibrary{positioning, arrows.meta}
  - \metroset{block=fill}
---

```{r install-packages, include=FALSE}
if (!length(getOption("repos")) || getOption("repos")["CRAN"] == "@CRAN@") {
  options(repos = c(CRAN = "https://cloud.r-project.org"))
}
required_packages <- c(
  "survival", "survminer", "dplyr", "knitr",
  "broom", "ggplot2", "forestmodel", "kableExtra", "tibble"
)
missing <- required_packages[!sapply(required_packages, requireNamespace, quietly = TRUE)]
if (length(missing) > 0) install.packages(missing)
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.width  = 11,
  fig.height = 4
)

library(survival)
library(survminer)
library(dplyr)
library(tibble)
library(knitr)
library(kableExtra)
library(broom)
library(ggplot2)
library(forestmodel)
```

```{r analysis, include=FALSE}
source(here::here("R", "main_analysis.R"))
```

## The clinical question

\vspace{12pt}

\begin{center}
\begin{tikzpicture}[font=\small, >=Stealth]
  % timeline
  \draw[->, line width=1pt] (-0.5, 0) -- (10.5, 0);

  % top labels
  \node[blue!60!black, font=\bfseries] at (0, 1.0) {Diagnosis};
  \node[orange!90!black, font=\bfseries] at (8.5, 1.0) {Toxicity};

  % icons on timeline
  \node[circle, fill=blue!30, draw=blue!60, minimum size=11mm, font=\bfseries\small] (ct) at (0,0) {CT};
  \node[circle, fill=orange!70, draw=orange!90!black, minimum size=11mm, font=\bfseries\small] (tox) at (8.5, 0) {!};

  % cycles
  \node[circle, fill=gray!20, minimum size=7mm] at (2.5, 0) {1};
  \node[circle, fill=gray!20, minimum size=7mm] at (4.0, 0) {2};
  \node[circle, fill=gray!20, minimum size=7mm] at (5.5, 0) {3};
  \node[circle, fill=gray!20, minimum size=7mm] at (7.0, 0) {4};
  \node[gray, font=\itshape\small] at (4.75, -0.7) {chemotherapy cycles};

  % HU + association arrow below
  \node[blue!60!black, font=\bfseries] at (0, -1.6) {Liver HU};
  \draw[->, dashed, thick, gray] (1.1, -1.6) -- (8.0, -1.6);
  \node[gray, font=\itshape] at (4.55, -2.0) {is it associated with the timing of toxicity?};
\end{tikzpicture}
\end{center}

\vspace{12pt}

\centerline{\small Capecitabine and oxaliplatin cause toxicities that force dose reductions, delays, or stopping treatment.}
\centerline{\small Liver fat at diagnosis is a plausible candidate marker.}

## The data

\begin{columns}[T,onlytextwidth]
\begin{column}{0.55\textwidth}

\textbf{Cohort:} COLON study, prospective, 11 Dutch hospitals.

\vspace{6pt}

\textbf{Three endpoints, separate models:}
\begin{itemize}
  \setlength\itemsep{1pt}
  \item CAP: capecitabine toxicity
  \item OXA: oxaliplatin toxicity
  \item CAPOX: combined regimen toxicity
\end{itemize}

\vspace{6pt}

\textbf{Time:} chemotherapy cycles to first dose-limiting toxicity (or censored at last cycle).

\vspace{6pt}

\textbf{Adjusted for:} age, sex, BMI, smoking, treatment-specific starting dose.

\end{column}
\begin{column}{0.42\textwidth}

```{r summary-tbl}
tibble(
  Endpoint     = c("CAP", "OXA", "CAPOX"),
  N            = c(nrow(cap), nrow(oxa), nrow(capox)),
  Events       = c(sum(cap$event), sum(oxa$event), sum(capox$event)),
  `Event rate` = sprintf("%.0f%%", c(mean(cap$event), mean(oxa$event), mean(capox$event)) * 100)
) %>%
  kable(format = "latex", booktabs = TRUE, linesep = "") %>%
  kable_styling(font_size = 11, latex_options = "hold_position")
```

\end{column}
\end{columns}

## What makes this tricky

```{r hu-boxplot, fig.height=3.4}
ggplot(df, aes(reorder(hospital, liverHU_manual, FUN = median, na.rm = TRUE),
               liverHU_manual)) +
  geom_boxplot(fill = "#56B4E9", alpha = 0.7, outlier.size = 0.7) +
  labs(x = NULL, y = "Liver radiodensity (HU)",
       title = "Liver HU varies systematically across hospitals") +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```

11 hospitals means 11 different CT scanners. Liver HU differs significantly between centres (ANOVA, p `r format.pval(anova(lm(liverHU_manual ~ hospital, data = df))[["Pr(>F)"]][1], digits = 2, eps = 0.001)`), most likely because of scanner calibration. Toxicity rates also differ for OXA, but for CAP and CAPOX the variation is consistent with sampling noise.

So hospital looks like a confounder we should not just ignore.

## How we handle it

\begin{columns}[T,onlytextwidth]
\begin{column}{0.48\textwidth}

\textbf{Unstratified Cox}

$$h(t \mid \mathbf{x}) = h_0(t)\, e^{\beta\,\text{HU} + \boldsymbol{\gamma}^\top \mathbf{x}}$$

\vspace{4pt}

One baseline hazard for everyone. The estimate of $\beta$ mixes within-hospital and between-hospital differences in liver HU.

\end{column}
\begin{column}{0.48\textwidth}

\textbf{Stratified by hospital}

$$h_k(t \mid \mathbf{x}) = h_{0k}(t)\, e^{\beta\,\text{HU} + \boldsymbol{\gamma}^\top \mathbf{x}}$$

\vspace{4pt}

Each hospital $k$ has its own baseline. The estimate of $\beta$ comes from \textit{within-hospital} comparisons only.

\end{column}
\end{columns}

\vspace{14pt}

\begin{block}{What we did}
Fit both specifications for each endpoint and compare. The difference between the two estimates tells us how much the unstratified version is being affected by between-hospital differences.
\end{block}

## The main result

\begin{columns}[T,onlytextwidth]
\begin{column}{0.40\textwidth}

\vspace{16pt}

\textbf{CAP and OXA}

\small Neither specification reaches statistical significance.

\vspace{14pt}

\normalsize \textbf{CAPOX}

\small Unstratified: HR $\approx$ 1.00, no association.\\
Stratified: HR = 1.10 (1.00 -- 1.21), p = 0.044.

\vspace{14pt}

\normalsize \textit{For CAPOX, between-hospital differences in measured HU were hiding the within-hospital signal.}

\end{column}
\begin{column}{0.58\textwidth}

\vspace*{\fill}

```{r forest-strat-vs-unstrat, fig.height=4.5, fig.width=5.5, out.width="100%"}
make_row <- function(model, label, type, scale = 10) {
  cf <- coef(model)["liverHU_manual"]
  se <- sqrt(vcov(model)["liverHU_manual", "liverHU_manual"])
  data.frame(
    Endpoint = label, Type = type,
    HR       = exp(cf * scale),
    CI_low   = exp((cf - 1.96 * se) * scale),
    CI_high  = exp((cf + 1.96 * se) * scale)
  )
}

forest_df <- bind_rows(
  make_row(cox_cap_unstrat,   "CAP",   "Unstratified"),
  make_row(cox_cap,           "CAP",   "Stratified"),
  make_row(cox_oxa_unstrat,   "OXA",   "Unstratified"),
  make_row(cox_oxa,           "OXA",   "Stratified"),
  make_row(cox_capox_unstrat, "CAPOX", "Unstratified"),
  make_row(cox_capox,         "CAPOX", "Stratified")
)
forest_df$Endpoint <- factor(forest_df$Endpoint, levels = rev(c("CAP", "OXA", "CAPOX")))
forest_df$Type     <- factor(forest_df$Type,     levels = c("Unstratified", "Stratified"))

ggplot(forest_df, aes(x = HR, y = Endpoint, color = Type)) +
  geom_vline(xintercept = 1, linetype = "dashed", color = "gray60") +
  geom_pointrange(aes(xmin = CI_low, xmax = CI_high),
                  position = position_dodge(width = 0.55), size = 0.7, fatten = 3) +
  scale_color_manual(values = c("Unstratified" = "#999999", "Stratified" = "#D55E00")) +
  scale_x_continuous(breaks = c(0.9, 1.0, 1.1, 1.2)) +
  labs(x = "HR per 10 HU (95% CI)", y = NULL, color = NULL) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "top",
        legend.direction = "horizontal",
        legend.margin = margin(0, 0, 0, 0))
```

\vspace*{\fill}

\end{column}
\end{columns}

## How big is the effect, in context?

For CAPOX (the conventionally significant endpoint):

\vspace{6pt}

\begin{center}
\begin{tabular}{lcc}
\toprule
\textbf{Scale} & \textbf{HR} & \textbf{95\% CI} \\
\midrule
per 10 HU                            & `r sprintf("%.2f", exp(10 * coef(cox_capox)["liverHU_manual"]))` & `r sprintf("%.2f -- %.2f", exp(10 * (coef(cox_capox)["liverHU_manual"] - 1.96 * sqrt(vcov(cox_capox)["liverHU_manual","liverHU_manual"]))), exp(10 * (coef(cox_capox)["liverHU_manual"] + 1.96 * sqrt(vcov(cox_capox)["liverHU_manual","liverHU_manual"]))))` \\[3pt]
per 1 SD ($\approx$ `r round(sd(capox$liverHU_manual, na.rm = TRUE), 0)` HU) & `r sprintf("%.2f", exp(sd(capox$liverHU_manual, na.rm = TRUE) * coef(cox_capox)["liverHU_manual"]))` & `r sprintf("%.2f -- %.2f", exp(sd(capox$liverHU_manual, na.rm = TRUE) * (coef(cox_capox)["liverHU_manual"] - 1.96 * sqrt(vcov(cox_capox)["liverHU_manual","liverHU_manual"]))), exp(sd(capox$liverHU_manual, na.rm = TRUE) * (coef(cox_capox)["liverHU_manual"] + 1.96 * sqrt(vcov(cox_capox)["liverHU_manual","liverHU_manual"]))))` \\
\bottomrule
\end{tabular}
\end{center}

\vspace{12pt}

A patient with a fatty liver compared to one with a healthy liver, about one within-cohort SD apart, has roughly a \textbf{`r sprintf("%.0f\\%%", 100 * (exp(sd(capox$liverHU_manual, na.rm = TRUE) * coef(cox_capox)["liverHU_manual"]) - 1))` higher hazard} of CAPOX toxicity.

\vspace{8pt}

\begin{block}{Caveats}
The lower bound of the CI is right at 1, the cohort is observational, and we have not validated this in another dataset. The effect is real in this cohort, but it is not yet a clinical tool.
\end{block}

## Takeaways

\begin{enumerate}
  \setlength\itemsep{8pt}
  \item \textbf{Stratification really mattered for our results.}\\
  Without it we would have concluded \textit{no association} for CAPOX. Hospital-level CT calibration was hiding the within-hospital signal.

  \item \textbf{There is a small within-hospital effect for CAPOX.}\\
  HR $\approx$ 1.10 per 10 HU, or about 1.21 per SD. No effect for CAP or OXA.

  \item \textbf{Not a clinical tool yet.}\\
  Effect is modest, the CI is wide, and the design is observational. The transferable lesson is methodological: in multi-centre cohorts with measurement heterogeneity, the within-hospital contrast can look quite different from the pooled one.
\end{enumerate}

# Backup material

## Hospital toxicity rates: chi-square test

Tests whether observed event-rate differences across hospitals exceed binomial sampling noise around a common rate.

```{r chisq-table}
tibble(
  Endpoint = c("CAP", "OXA", "CAPOX"),
  N        = c(sum(table(cap$hospital, cap$event)),
               sum(table(oxa$hospital, oxa$event)),
               sum(table(capox$hospital, capox$event))),
  `Chi-sq` = c(round(unname(chi_cap$statistic),   2),
               round(unname(chi_oxa$statistic),   2),
               round(unname(chi_capox$statistic), 2)),
  df       = c(unname(chi_cap$parameter),
               unname(chi_oxa$parameter),
               unname(chi_capox$parameter)),
  `p-value` = sapply(c(chi_cap$p.value, chi_oxa$p.value, chi_capox$p.value),
                     function(p) ifelse(p < 0.001, "<0.001", sprintf("%.3f", p)))
) %>%
  kable(format = "latex", booktabs = TRUE, linesep = "") %>%
  kable_styling(font_size = 11, latex_options = "hold_position")
```

\vspace{8pt}

\begin{itemize}
  \setlength\itemsep{2pt}
  \item Smallest hospital $\times$ event cell count is 7, so the asymptotic chi-square is reliable (no need for Monte Carlo simulation).
  \item For OXA the differences are real. For CAP and CAPOX the observed range is consistent with sampling noise around a common rate.
  \item Stratification is still justified by the liver HU heterogeneity (ANOVA), regardless of this test.
\end{itemize}

## Full covariate model: CAP

```{r forest-cap-full, fig.height=4.2}
forest_model(cox_cap)
```

For CAP, none of the covariates reach statistical significance. The CAP model is generally less informative, suggesting that capecitabine-related toxicity is less strongly tied to the patient characteristics included here.

## Full covariate model: OXA

```{r forest-oxa-full, fig.height=4.2}
forest_model(cox_oxa)
```

For OXA, age, sex (male less affected), smoking status, and oxaliplatin starting dose are all significantly associated with toxicity. Liver HU sits just outside conventional significance (p = 0.060).

## Full covariate model: CAPOX

```{r forest-capox-full, fig.height=4.2}
forest_model(cox_capox)
```

For CAPOX, age, sex, and smoking show consistent associations with toxicity, similar to the OXA model. Liver HU reaches conventional significance (p = 0.044).

## Proportional hazards check

Schoenfeld residual test, GLOBAL row per model. A non-significant p means no evidence the proportional-hazards assumption is violated.

```{r ph-table}
ph_global <- data.frame(
  Endpoint = c("CAP", "OXA", "CAPOX"),
  `Chi-sq` = sapply(list(ph_cap, ph_oxa, ph_capox),
                    function(x) round(x$table["GLOBAL", "chisq"], 2)),
  df       = sapply(list(ph_cap, ph_oxa, ph_capox),
                    function(x) x$table["GLOBAL", "df"]),
  `p-value` = sapply(list(ph_cap, ph_oxa, ph_capox),
                     function(x) {
                       p <- x$table["GLOBAL", "p"]
                       ifelse(p < 0.001, "<0.001", sprintf("%.3f", p))
                     }),
  check.names = FALSE
)

kable(ph_global, format = "latex", booktabs = TRUE, linesep = "",
      row.names = FALSE) %>%
  kable_styling(font_size = 11, latex_options = "hold_position")
```

\vspace{8pt}

All three GLOBAL p-values are well above 0.05. Per-covariate Schoenfeld results (also non-significant) are in the analyst notebook.

## Kaplan-Meier curves

```{r km-curves, fig.height=4.0}
fit_cap   <- survfit(Surv(time, event) ~ 1, data = cap)
fit_oxa   <- survfit(Surv(time, event) ~ 1, data = oxa)
fit_capox <- survfit(Surv(time, event) ~ 1, data = capox)

p_cap   <- ggsurvplot(fit_cap,   data = cap,   conf.int = TRUE, title = "CAP",
                      xlab = "Cycles", ylab = "Toxicity-free",
                      ggtheme = theme_minimal(base_size = 10))
p_oxa   <- ggsurvplot(fit_oxa,   data = oxa,   conf.int = TRUE, title = "OXA",
                      xlab = "Cycles", ylab = "Toxicity-free",
                      ggtheme = theme_minimal(base_size = 10))
p_capox <- ggsurvplot(fit_capox, data = capox, conf.int = TRUE, title = "CAPOX",
                      xlab = "Cycles", ylab = "Toxicity-free",
                      ggtheme = theme_minimal(base_size = 10))

arrange_ggsurvplots(list(p_cap, p_oxa, p_capox), ncol = 3, nrow = 1)
```

Most events occur in the first few cycles. The probability of remaining toxicity-free drops sharply early on, then stabilises as fewer patients remain at risk.

## Sample sizes per hospital (CAP)

```{r per-hospital}
df %>%
  filter(!is.na(TIMT_CAP2)) %>%
  group_by(Hospital = hospital) %>%
  summarise(
    N        = n(),
    Events   = sum(TIMT_CAP2 == 1),
    `Rate (%)` = round(100 * mean(TIMT_CAP2 == 1, na.rm = TRUE), 0),
    .groups = "drop"
  ) %>%
  arrange(desc(N)) %>%
  kable(format = "latex", booktabs = TRUE, linesep = "") %>%
  kable_styling(font_size = 9, latex_options = "hold_position")
```

\vspace{6pt}

Hospital sizes range from very small to ~80 patients. The smallest hospitals contribute little to the within-hospital comparisons, which is appropriate: stratification gives them their own baseline rather than letting them drag a pooled estimate around.
