---
title: "Does Liver Radiodensity Predict Chemotherapy Toxicity?"
subtitle: "Time-to-event analysis across CAP, OXA, and CAPOX"
author: "Noah Bos & Abdullah Alam"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    number_sections: true
    df_print: paged
---

```{r install-packages, include=FALSE}
required_packages <- c(
  "survival", "survminer", "dplyr", "knitr",
  "broom", "ggplot2", "forestmodel", "kableExtra", "tibble"
)
missing <- required_packages[!sapply(required_packages, requireNamespace, quietly = TRUE)]
if (length(missing) > 0) install.packages(missing)
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.width  = 9,
  fig.height = 5
)

library(survival)
library(survminer)
library(dplyr)
library(tibble)
library(knitr)
library(kableExtra)
library(broom)
library(ggplot2)
library(forestmodel)
```

```{r data-load, include=FALSE}
load("data/Dataset_liverfat.RData")
df <- dataset_leiden
```

```{r data-prep, include=FALSE}
# Constructs a complete-case analysis dataset for a given toxicity endpoint.
# Each endpoint has its own risk set (not all patients received all drugs),
# its own time variable, and its own dose covariate(s).
# complete.cases() ensures all model covariates are non-missing.
# This means N here (314/289/339) is smaller than the raw eligible N
# (344/311/359) — the difference is patients missing at least one covariate.

make_endpoint <- function(data, endpoint) {

  if (endpoint == "CAP") {
    event <- data$TIMT_CAP2
    time  <- data$cycle_till_timt
    dose  <- data$starting_doseCAP
  }

  if (endpoint == "OXA") {
    event <- data$TIMT_OXA2
    time  <- data$cycle_till_timtOXA
    dose  <- data$starting_doseOXA
  }

  if (endpoint == "CAPOX") {
    event   <- data$TIMT_CAPOX2
    time    <- data$cycle_till_timtCAPOX
    doseCAP <- data$starting_doseCAP
    doseOXA <- data$starting_doseOXA
  }

  out <- data.frame(
    time, event,
    liverHU_manual = data$liverHU_manual,
    age      = data$age,
    sex      = data$sex,
    BBMI     = data$BBMI,
    BSMOKER  = data$BSMOKER,
    hospital = data$hospital
  )

  if (endpoint == "CAP" || endpoint == "OXA") out$starting_dose <- dose
  if (endpoint == "CAPOX") { out$doseCAP <- doseCAP; out$doseOXA <- doseOXA }

  out[complete.cases(out), ]
}

cap   <- make_endpoint(df, "CAP")
oxa   <- make_endpoint(df, "OXA")
capox <- make_endpoint(df, "CAPOX")
```

---

# Study Overview

## Research question

Does **liver radiodensity** (measured in Hounsfield Units, HU) predict
**time to dose-limiting toxicity (DLT)** in patients receiving chemotherapy?

Liver radiodensity is a CT-derived measure of liver fat content. Lower HU
values indicate more fat (hepatic steatosis), which may affect drug metabolism
and therefore toxicity risk.

## Why time-to-event analysis?

Patients are followed over multiple chemotherapy cycles and may stop
(be censored) for reasons other than toxicity — e.g. disease progression or
end of follow-up. Standard logistic regression would ignore this structure.
Cox proportional hazards models correctly handle both the timing and the
censoring of events.

## Three separate endpoints

Not all patients receive both drugs, so each endpoint has its own risk set:

| Endpoint | Drug(s) | Dose covariate(s) |
|---|---|---|
| CAP | Capecitabine only | starting_doseCAP |
| OXA | Oxaliplatin only | starting_doseOXA |
| CAPOX | Both combined | starting_doseCAP + starting_doseOXA |

---

# Descriptive Statistics

## Sample sizes per endpoint

The table below shows the number of patients, events (DLT occurred), and
censored observations in each analysis dataset. These are **complete-case
counts** — patients missing any model covariate are excluded.

```{r endpoint-summary}
endpoint_summary <- tibble(
  Endpoint      = c("CAP", "OXA", "CAPOX"),
  N             = c(nrow(cap),          nrow(oxa),          nrow(capox)),
  Events        = c(sum(cap$event),     sum(oxa$event),     sum(capox$event)),
  Censored      = c(sum(cap$event == 0), sum(oxa$event == 0), sum(capox$event == 0)),
  `Event rate`  = paste0(round(100 * c(
    sum(cap$event)   / nrow(cap),
    sum(oxa$event)   / nrow(oxa),
    sum(capox$event) / nrow(capox)
  ), 1), "%")
)

kable(endpoint_summary, caption = "Complete-case analysis sample sizes") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

## Liver radiodensity across hospitals

The boxplot below shows the distribution of liver HU per hospital. Systematic
differences exist between hospitals, which means hospital is a potential
**confounder**: it affects both HU (through patient case-mix) and toxicity
rates (through treatment practices). This motivates stratifying by hospital
in the Cox models.

```{r hu-by-hospital, fig.height=4}
ggplot(df, aes(x = reorder(hospital, liverHU_manual, median), y = liverHU_manual)) +
  geom_boxplot(fill = "#AEC6E8", color = "#2166AC") +
  labs(
    title = "Liver radiodensity (HU) by hospital",
    x     = "Hospital (ordered by median HU)",
    y     = "Liver HU"
  ) +
  theme_minimal(base_size = 12)
```

---

# Kaplan-Meier Curves

Kaplan-Meier curves show the **unadjusted** probability of remaining
toxicity-free over chemotherapy cycles. These are descriptive only — they do
not adjust for any covariates or hospital differences.

The risk table below each curve shows how many patients remained under
observation at each cycle. As the number at risk drops, the survival estimate
becomes less reliable (wider confidence intervals).

## CAP

```{r km-cap, fig.height=6}
ggsurvplot(
  survfit(Surv(time, event) ~ 1, data = cap),
  data       = cap,
  risk.table = TRUE,
  conf.int   = TRUE,
  title      = "Time to toxicity: CAP",
  xlab       = "Chemotherapy cycles",
  ylab       = "Probability toxicity-free",
  ggtheme    = theme_minimal(base_size = 12)
)
```

## OXA

```{r km-oxa, fig.height=6}
ggsurvplot(
  survfit(Surv(time, event) ~ 1, data = oxa),
  data       = oxa,
  risk.table = TRUE,
  conf.int   = TRUE,
  title      = "Time to toxicity: OXA",
  xlab       = "Chemotherapy cycles",
  ylab       = "Probability toxicity-free",
  ggtheme    = theme_minimal(base_size = 12)
)
```

## CAPOX

```{r km-capox, fig.height=6}
ggsurvplot(
  survfit(Surv(time, event) ~ 1, data = capox),
  data       = capox,
  risk.table = TRUE,
  conf.int   = TRUE,
  title      = "Time to toxicity: CAPOX",
  xlab       = "Chemotherapy cycles",
  ylab       = "Probability toxicity-free",
  ggtheme    = theme_minimal(base_size = 12)
)
```

## KM curves by hospital (CAPOX)

The plot below shows KM curves split by hospital for the CAPOX endpoint.
Visible separation between hospitals confirms that baseline toxicity risk
differs across centres — another reason to stratify.

```{r km-by-hospital, fig.height=6}
ggsurvplot(
  survfit(Surv(time, event) ~ hospital, data = capox),
  data          = capox,
  censor        = FALSE,
  legend.title  = "Hospital",
  legend        = "right",
  title         = "Time to CAPOX toxicity by hospital (unadjusted)",
  xlab          = "Chemotherapy cycles",
  ylab          = "Probability toxicity-free",
  ggtheme       = theme_minimal(base_size = 12)
)
```

---

# Cox Proportional Hazards Models

## Model specification

Each endpoint is modelled with a **stratified Cox model**:

- **Fixed effects:** liver HU, age, sex, BMI, smoking status, starting dose
- **Stratification:** hospital — each hospital gets its own baseline hazard
  function, so the model estimates the HU effect *within* hospitals rather
  than *between* them

Stratifying by hospital (rather than adjusting for it as a fixed effect)
is more flexible: it makes no assumption about the shape of the baseline
hazard, only that the effect of liver HU is the same across hospitals.

```{r fit-cox-models}
cox_cap <- coxph(
  Surv(time, event) ~ liverHU_manual + age + sex + BBMI +
    BSMOKER + starting_dose + strata(hospital),
  data = cap
)

cox_oxa <- coxph(
  Surv(time, event) ~ liverHU_manual + age + sex + BBMI +
    BSMOKER + starting_dose + strata(hospital),
  data = oxa
)

cox_capox <- coxph(
  Surv(time, event) ~ liverHU_manual + age + sex + BBMI +
    BSMOKER + doseCAP + doseOXA + strata(hospital),
  data = capox
)
```

## Hazard ratio table

The table below shows adjusted hazard ratios (HR) for all covariates across
the three models. HR > 1 means higher toxicity risk; HR < 1 means lower risk.
The 95% CI and p-value indicate statistical uncertainty and significance.

**Focus on the `liverHU_manual` row** — that is the primary exposure of interest.

```{r hr-table}
hr_table <- bind_rows(
  tidy(cox_cap,   exponentiate = TRUE, conf.int = TRUE) %>% mutate(Endpoint = "CAP"),
  tidy(cox_oxa,   exponentiate = TRUE, conf.int = TRUE) %>% mutate(Endpoint = "OXA"),
  tidy(cox_capox, exponentiate = TRUE, conf.int = TRUE) %>% mutate(Endpoint = "CAPOX")
) %>%
  mutate(
    term     = gsub("BSMOKER1", "Smoker", term),
    term     = gsub("sex1",     "Sex (male)", term),
    sig      = ifelse(p.value < 0.05, "*", "")
  ) %>%
  select(Endpoint, Term = term, HR = estimate, `CI low` = conf.low,
         `CI high` = conf.high, `p-value` = p.value, ` ` = sig)

kable(hr_table, digits = 3,
      caption = "Adjusted hazard ratios from stratified Cox models (* p < 0.05)") %>%
  kable_styling(bootstrap_options = c("striped", "hover")) %>%
  row_spec(which(hr_table$Term == "liverHU_manual"), bold = TRUE, background = "#EFF3FF")
```

## Forest plots

Forest plots visualise the same hazard ratios graphically. Each point is the
HR estimate; the horizontal bar is the 95% CI. The dashed line at 1.0 is the
null (no effect). Points to the right of 1 indicate increased toxicity risk.

### CAP

```{r forest-cap}
forest_model(cox_cap)
```

### OXA

```{r forest-oxa}
forest_model(cox_oxa)
```

### CAPOX

```{r forest-capox}
forest_model(cox_capox)
```

---

# Model Diagnostics: Proportional Hazards Assumption

The Cox model assumes that hazard ratios are **constant over time** (the
proportional hazards assumption). This is tested using Schoenfeld residuals.

- A **non-significant p-value** (> 0.05) means no strong evidence of violation
- A **significant p-value** for a covariate suggests its effect changes over
  time, which would invalidate the HR interpretation for that covariate

```{r ph-tests}
ph_cap   <- cox.zph(cox_cap)
ph_oxa   <- cox.zph(cox_oxa)
ph_capox <- cox.zph(cox_capox)
```

## CAP

```{r ph-table-cap}
kable(as.data.frame(ph_cap$table), digits = 3,
      caption = "Schoenfeld residual test: CAP") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

## OXA

```{r ph-table-oxa}
kable(as.data.frame(ph_oxa$table), digits = 3,
      caption = "Schoenfeld residual test: OXA") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

## CAPOX

```{r ph-table-capox}
kable(as.data.frame(ph_capox$table), digits = 3,
      caption = "Schoenfeld residual test: CAPOX") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

---

# Conclusions

- Liver radiodensity (HU) shows **no consistent or statistically significant
  association** with time to dose-limiting toxicity across any of the three
  endpoints
- Hazard ratios for liver HU are close to 1 with wide confidence intervals
- Other covariates — particularly **starting dose** — show stronger
  associations with toxicity risk
- Hospital stratification was appropriate: KM curves by hospital showed
  meaningful variation in baseline toxicity rates
- The proportional hazards assumption was not violated in any model

## Clinical interpretation

A hazard ratio of ~0.99 per 1 HU corresponds to roughly 0.90 per 10 HU —
a 10% reduction in hazard. Even if real, this effect size is unlikely to
change clinical decisions. Liver radiodensity alone does not appear to add
meaningful predictive value beyond standard clinical variables.

## Future directions

This analysis modelled liver HU continuously. A possible follow-up would be
to examine whether toxicity risk differs at **clinically meaningful thresholds**
(e.g. HU < 40, indicating moderate–severe steatosis) rather than across the
full continuous range.
