---
title: "Does Liver Radiodensity Predict Chemotherapy Toxicity?"
subtitle: "Time-to-event analysis across CAP, OXA, and CAPOX"
author: "Noah Bos & Abdullah Alam"
date: "`r Sys.Date()`"

output:
  beamer_presentation:
    theme: "Madrid"
---

```{r install-packages, include=FALSE}
required_packages <- c(
  "survival", "survminer", "dplyr", "knitr",
  "broom", "ggplot2", "forestmodel", "kableExtra", "tibble"
)
missing <- required_packages[!sapply(required_packages, requireNamespace, quietly = TRUE)]
if (length(missing) > 0) install.packages(missing)
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 5
)

library(survival)
library(survminer)
library(dplyr)
library(tibble)
library(knitr)
library(kableExtra)
library(broom)
library(ggplot2)
library(forestmodel)
```

```{r}
load("data/Dataset_liverfat.RData")
df = dataset_leiden
```

```{r data-prep, include=FALSE}
## FIX: make_endpoint moved here so analysis datasets (cap/oxa/capox) exist
## before the summary table chunk. Previously the function was defined at
## line ~85, AFTER the summary table — so the table counted raw N
## (non-missing event only: 344/311/359) while the KM curves used
## complete-case N (314/289/339). The gap of 30/22/20 patients per endpoint
## consists of individuals who have a non-missing event indicator but are
## missing at least one model covariate (liverHU_manual, age, sex, BBMI,
## BSMOKER, hospital, or starting dose). Both numbers are correct for their
## respective purpose; this fix makes the summary table show the analysis N.

make_endpoint <- function(data, endpoint){

  if(endpoint=="CAP"){
    event <- data$TIMT_CAP2
    time  <- data$cycle_till_timt
    dose  <- data$starting_doseCAP
  }

  if(endpoint=="OXA"){
    event <- data$TIMT_OXA2
    time  <- data$cycle_till_timtOXA
    dose  <- data$starting_doseOXA
  }

  if(endpoint=="CAPOX"){
    event   <- data$TIMT_CAPOX2
    time    <- data$cycle_till_timtCAPOX
    ## FIX: added both dose variables for CAPOX; previously neither was
    ## captured here, so they were silently absent from the dataset and
    ## excluded from the Cox model.
    doseCAP <- data$starting_doseCAP
    doseOXA <- data$starting_doseOXA
  }

  out <- data.frame(
    time, event,
    liverHU_manual = data$liverHU_manual,
    age      = data$age,
    sex      = data$sex,
    BBMI     = data$BBMI,
    BSMOKER  = data$BSMOKER,
    hospital = data$hospital
  )

  if(endpoint == "CAP" || endpoint == "OXA"){
    out$starting_dose <- dose
  }

  ## FIX: attach CAPOX dose variables to the output data frame
  if(endpoint == "CAPOX"){
    out$doseCAP <- doseCAP
    out$doseOXA <- doseOXA
  }

  out <- out[complete.cases(out), ]
  out
}

cap   <- make_endpoint(df, "CAP")
oxa   <- make_endpoint(df, "OXA")
capox <- make_endpoint(df, "CAPOX")
```

# Study Objective

- Assess whether liver radiodensity (HU) predicts time to dose-limiting toxicity  
- Outcome: number of cycles until toxicity occurs  
- Adjusted for patient characteristics and hospital differences  

**Key takeaway:** Liver HU is evaluated for its added predictive value in clinical decision-making.

# Toxicity Endpoints Analyzed Separately

- CAP toxicity  
- OXA toxicity  
- CAPOX toxicity  

**Key takeaway:** Each treatment is evaluated using a separate time-to-event model.

# Observed Toxicity Rates Differ by Treatment

```{r}
## FIX: count N from the complete-case analysis datasets so numbers match
## the KM curves. Raw eligible N (non-missing event only) was 344/311/359;
## analysis N after dropping rows with any missing covariate is 314/289/339.
endpoint_summary <- tibble(
  Endpoint = c("CAP", "OXA", "CAPOX"),
  N        = c(nrow(cap),   nrow(oxa),   nrow(capox)),
  Events   = c(sum(cap$event),  sum(oxa$event),  sum(capox$event)),
  Censored = c(sum(cap$event==0), sum(oxa$event==0), sum(capox$event==0))
)

kable(endpoint_summary)
```

**Key takeaway:** Event rates differ across treatments, with sufficient events to support survival analysis.

# Liver Radiodensity Varies Across Hospitals

```{r}
ggplot(df, aes(hospital, liverHU_manual)) +
  geom_boxplot() +
  theme_minimal()
```

**Key takeaway:** Liver HU distributions differ between hospitals, suggesting baseline differences that are accounted for through stratification.


```{r include=FALSE}
## FIX: make_endpoint and dataset creation moved to the data-prep chunk above.
## This chunk is retained as a placeholder to avoid slide numbering shifts.
```

# Time to Toxicity: CAP

```{r}
ggsurvplot(
survfit(Surv(time,event)~1,data=cap),
risk.table=T
)
```
**Key takeaway:** The probability of remaining toxicity-free declines over cycles, with most events occurring early in treatment.


# Time to Toxicity: OXA

```{r}
ggsurvplot(
survfit(Surv(time,event)~1,data=oxa),
risk.table=T
)
```
**Key takeaway:** Toxicity risk is highest in early cycles. Estimates in later cycles are less reliable due to few patients remaining at risk.


# Time to Toxicity: CAPOX

```{r}
ggsurvplot(
survfit(Surv(time,event)~1,data=capox),
risk.table=T
)
```
**Key takeaway:** A similar pattern is observed, with most toxicities occurring early and increasing uncertainty in later cycles.

# Kaplan-Meijer curves by hospital (capox as example)

```{r}
fit_hosp <- survfit(Surv(time, event) ~ hospital, data = capox)

ggsurvplot(
  fit_hosp,
  data = capox,
  censor = FALSE,
  legend.title = "Hospital",
  legend = "right",
  ggtheme = theme_minimal(base_size = 16)
)

```

- toxicity risk varies across hospitals
- differences likely reflect patient mix and clinical practice

# Why Use Cox Models?

```{r}
cox_cap <- coxph(
Surv(time,event)~liverHU_manual+age+sex+BBMI+
BSMOKER+starting_dose+strata(hospital),
data=cap)

cox_oxa <- coxph(
Surv(time,event)~liverHU_manual+age+sex+BBMI+
BSMOKER+starting_dose+strata(hospital),
data=oxa)

## FIX: added doseCAP + doseOXA to CAPOX model; both doses were present in
## the older analysis files but had been accidentally omitted here.
cox_capox <- coxph(
Surv(time,event)~liverHU_manual+age+sex+BBMI+
BSMOKER+doseCAP+doseOXA+strata(hospital),
data=capox)
```

- Kaplan-Meijer curves are unadjusted and do not account for hospital differences

- Cox models are stratified by hospital  
- This allows each hospital to have its own baseline hazard  

**Why this matters:**

- Accounts for differences in patient populations, practices, and measurement  
- Ensures estimated effects (e.g., liver HU) are not confounded by hospital differences  

**Key takeaway:** Results reflect within-hospital comparisons, not differences between hospitals.


# Association Between Liver HU and Toxicity

```{r }
hr_table <- bind_rows(
  tidy(cox_cap, exponentiate = TRUE, conf.int = TRUE) %>% mutate(Model="CAP"),
  tidy(cox_oxa, exponentiate = TRUE, conf.int = TRUE) %>% mutate(Model="OXA"),
  tidy(cox_capox, exponentiate = TRUE, conf.int = TRUE) %>% mutate(Model="CAPOX")
)

kable(hr_table, digits = 2) %>%
  kableExtra::kable_styling(font_size = 7)
```

**Key takeaway:** Hazard ratios for liver HU are close to 1 across all endpoints, with no statistically significant associations observed.

# CAP: No Meaningful Association with Liver HU

```{r}
forest_model(cox_cap)
```

**Key takeaway:** The estimated effect of liver HU is close to null with no statistically significant association.

# OXA: Small Trend, Not Statistically significant

```{r}
forest_model(cox_oxa)
```

**Key takeaway:** A slight association is observed, but the effect size is small and not significant


# CAPOX: No Consistent Association Observed

```{r}
forest_model(cox_capox)
```

**Key takeaway:** Estimates are close to null, with no clear evidence of an association.

# Model Assumptions Hold

- No evidence of violation of the proportional hazards assumption  
- Hazard ratios appear stable over time  

```{r}
## FIX: Schoenfeld residual tests added for CAP and OXA; previously only
## CAPOX was tested.
ph_cap   <- cox.zph(cox_cap)
ph_oxa   <- cox.zph(cox_oxa)
ph_capox <- cox.zph(cox_capox)

kable(as.data.frame(ph_cap$table),   caption = "PH test: CAP")
kable(as.data.frame(ph_oxa$table),   caption = "PH test: OXA")
kable(as.data.frame(ph_capox$table), caption = "PH test: CAPOX")
```

**Key takeaway:** The Cox model assumptions are satisfied, supporting the validity of the results.

# Overall Interpretation

- Liver radiodensity shows no consistent or clinically meaningful association with time to toxicity  
- Effect sizes are small and not statistically significant across endpoints  
- Other clinical variables (e.g., age, smoking, dose) show stronger associations  

**Key takeaway:** Liver radiodensity does not appear to add meaningful predictive value beyond standard clinical variables even after accounting for hospital effects

# Future study

No clear association was observed when liver radiodensity was modeled continuously across its full range. A possible follow-up analysis would be to assess whether toxicity risk differs only at clinically meaningful thresholds or extreme radiodensity ranges.
